---
title: "Advanced Normalization Tools Quick Reference"
author: "Brian B. Avants"
date: "July 15, 2015"
output: rmarkdown::tufte_handout
---

```{r, echo=FALSE}
library(ggplot2)
library(ANTsR)
```

# Introduction

\newthought{Advanced Normalization Tools}[^ants] is software for biomedical image analysis with a focus on registration, segmentation, geometric quantification and statistics. 
Statistical methods are available in *ANTsR* ^[http://stnava.github.io/ANTsR/] which tightly integrates *ANTs* with the *R* statistical computing language.  This document briefly highlights *ANTs* features.

# Provenance and Testing

Much core functionality in *ANTs* lives in ITK, a project to which we contribute regularly.  This core is tested on many platforms via `ctest` and the results are on the ITK dashboard^[https://open.cdash.org/index.php?project=Insight].
*ANTs* is also tested with every commit via codeship ^[https://codeship.com] and Travis ^[https://travis-ci.org/stnava/ANTs].  Many (not all) *ANTs* programs support `programName --version`.  Our github commit hashes give the best way to track code versions by identifying core and dependency versions.  See the *ANTs* website for current testing results and **installation** instructions.

# In conjunction with other analysis systems

ANTs and ITK work synergistically within a well-defined I/O and world
coordinate system.  Using FSL, SPM or other pre-processing in 
conjunction with ITK and *ANTs* must be done with extreme care. The 
physical spaces (or interpretation of image headers) may not be the same in different systems.  Such inconsistencies may lead to severe misinterpretation of results. \marginnote{WARNING!}

# Data types

## Images
Images are the core data type in ANTs. Valid extensions are determined by ITK ^[www.itk.org] image input/output libraries ^[NiftiImageIO, NrrdImageIO, GiplImageIO, HDF5ImageIO, JPEGImageIO, GDCMImageIO, BMPImageIO, LSMImageIO, PNGImageIO, TIFFImageIO, VTKImageIO, StimulateImageIO, BIORadImageIO,
MetaImageIO, MRCImageIO, GE4ImageIO, GE5ImageIO, MGHImageIO].  ITK images can be of arbitrary dimensionality and pixel type, but in *ANTs* we instantiate `float` pixel type in 2, 3 or 4 dimensions.


## CSV files and point sets

CSV (comma separated value) files are useful for storing tabular data.  We may also use them to store point sets.  It is critical when using point sets with *ANTs* to ensure that the physical space of the points matches that of the images.  It is best to start with our standard example of registering images and applying the results to points ^[https://github.com/stnava/chicken].  Point sets may also be stored in the binary format provided by the Meta I/O library ^[http://www.itk.org/Wiki/ITK/MetaIO/Documentation].  This format allows much faster I/O for large datasets.


# Registration

The most well-known components of *ANTs* relate to mapping between coordinate systems ^[Registration: targeting of points in one coordinate system onto another].  Briefly, these maps may relate to:

* statistical image similarity: image difference, correlation or mutual information

* translation, rotation, affine, B-spline or various diffeomorphic models ^[Diffeomorphism: differentiable map with differentiable inverse]

* spatial or spatiotemporal maps

* dense or sparse transformation models, the latter being similar to SIFT or HOGG

* applying composite transformations to images, labels or point sets.

These components all exist within an integrated framework which is nearly always capable of incorporating many features or applying to multiple different modalities.^[http://journal.frontiersin.org/article/10.3389/fninf.2014.00044/abstract]

## Evaluation history

ANTs became well known because it performed well in a variety of open competitions related to image registration.  Most of these successes occurred in the days before the community had many such competitions. ^[Klein 2009, Murphy 2011, SATA 201*, BRATS 2013, etc.]  Highlights include:  

* finishing in the top rank in the Klein 2009 evaluation on brain MRI

* finishing in the top rank in the Murphy 2011 lung CT evaluation

* top SATA 2012 and 2013 finishers used ANTs

* performing well in a cardiac motion estimation competition

* well-known robust performance on large datasets 

Although *ANTs* has often performed well without using domain knowledge, it is still valuable to use problem-specific solutions when feasible.

## Quick start

ANTsR gives some quick registration options.  One can achieve similar performance with `antsRegistrationSyNQuick.sh`.

```{r reg, fig.cap='quick parameter free registration',fig.fullwidth = FALSE,message=FALSE,warning=FALSE}
fi <- antsImageRead(getANTsRData("r16") )
mi <- antsImageRead(getANTsRData("r64") )
mytx <- antsRegistration(fixed=fi, moving=mi, typeofTransform = c('SyN') )
mywarpedimage <- antsApplyTransforms( fixed=fi, moving=mi,
  transformlist=mytx$fwdtransforms )
```

```{r reg0, fig.cap='initial relative image positions',fig.fullwidth = FALSE,message=FALSE,warning=FALSE}
plot(fi, mi %>% iMath("Canny",1,5,12) )
```

```{r reg1, fig.cap='quick parameter free registration',fig.fullwidth = FALSE,message=FALSE,warning=FALSE}
plot(fi, mywarpedimage %>% iMath("Canny",1,5,12) )
```

## Low-dimensional

translation, rigid, affine - optional multi-start exploration of the transformation space (`antsAI`)

## High-dimensional

transformations with many parameters - primarily `SyN` and `B-spline SyN` as well as time-varying diffeomorphic models.

## Parameter choices and testing

We wrote a paper that details procedures for evaluating analysis software ^[http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3766821/].  We wrote this paper in order to help those who want to use or evaluate *ANTs* make more informed choices on how to proceed.  Briefly: (1) ask developers questions; (2) leverage biologically motivated testing metrics that are independent of registration.

```{r, fig.width = 10, fig.height = 2, fig.fullwidth = TRUE, fig.cap = "Full width figure",eval=FALSE,echo=FALSE}
# You can also include \LaTeX\ equations in the margin by explicitly invoking the `marginfigure` environment.

\begin{marginfigure}
$$\frac{d}{dx}\left( \int_{0}^{x} f(u)\,du\right)=f(x).$$
\caption{An equation}
\end{marginfigure}

# You can arrange for figures to span across the entire page by using the `fig.fullwidth` chunk option. 
qplot(wt, mpg, data=mtcars, colour=factor(cyl))
```



```{r, fig.cap = "Another figure",echo=FALSE,eval=FALSE}
# Note the use of the `fig.width` and `fig.height` chunk options to establish the proportions of the figure. Full width figures look much better if their height is minimized.
# Besides margin and full width figures, you can of course also include figures constrained to the main column.
qplot(factor(cyl), mpg, data = mtcars, geom = "boxplot")
```

# Segmentation, geometry and labeling

This category of methods relates to labeling and quantifying images.  These methods may require data of reasonable quality to perform well. \marginnote{WARNING!}  

## N4

Nick's N3 ^[http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3071855/] improves the original N3 inhomogeneity correction method.

```{r n4}
ficorrupt = as.antsImage( as.array( fi ) * 1:length(fi[fi>=0]) )
fifixed = n4BiasFieldCorrection( ficorrupt )
```

```{r n4a, fig.cap='corrupted image',fig.fullwidth = FALSE,message=FALSE,warning=FALSE}
plot( ficorrupt )
```

```{r n4b, fig.cap='n4 repair',fig.fullwidth = FALSE,message=FALSE,warning=FALSE}
plot( fifixed )
```


## Atropos

Expecation maximization segmentation with a variety of likelihood models and initialization strategies.  Incorporates multiple modalities and allows control of prior strength.  The finite mixture modeling (FMM) segmentation approach is the most popular. Prior constraints include the specification of a prior label image, prior probability images (one for each class), and/or an MRF prior to enforce spatial smoothing of the labels. 

```{r atropos}
img <- antsImageRead( getANTsRData("r16")  )
mask <- getMask(img)
segs1 <- kmeansSegmentation( img, 3, mask )

# Use probabilities from k-means seg as priors
feats <- list(img, iMath(img,"Laplacian"), iMath(img,"Grad") )
segs2 <- atropos( d = 2, a = feats, m = '[0.2,1x1]',
   c = '[2,0]',  i = segs1$probabilityimages, x = mask )
```



```{r atropos_0, fig.cap='kmeans result',fig.fullwidth = FALSE,message=FALSE,warning=FALSE}
plot( fi, segs1$segmentation )
```


```{r atropos_1, fig.cap='prior probability result with 2 feature channels',fig.fullwidth = FALSE,message=FALSE,warning=FALSE}
plot( fi, segs2$segmentation )
```

## Weingarten map-based surface curvature and area

The shape operator provides a beautiful way to compute the mean or Gaussian curvature (or related values) in any three-dimensional image ^[http://www.ncbi.nlm.nih.gov/pubmed/15344450].  The *ANTs* program that implements this is called `SurfaceCurvature`.

```{r shapeop}
fi <- antsImageRead( getANTsRData( "mni" ) )
fiseg = kmeansSegmentation( fi, 3 )
fik <- weingartenImageCurvature( fi )
```


```{r shapeopViz, fig.cap='shape operator in 3D',fig.fullwidth = FALSE,message=FALSE,warning=FALSE}
plot(  fik, axis=3 )
```

```{r shapeopViz2, fig.cap='shape operator selects sulci',fig.fullwidth = FALSE,message=FALSE,warning=FALSE}
fisulc = antsImageClone( fik ) * 0
selector = ( fiseg$segmentation == 2 & fik < 127.999 )
fisulc[  selector ] = fik[ selector ]
plot(  fi, fisulc, axis=3 )
```

# Cortical thickness pipeline

You can use `antsCorticalThickness.sh` to reproduce our lifespan thickness analysis from the Freesurfer evaluation paper ^[http://www.ncbi.nlm.nih.gov/pubmed/24879923].

# Multiple modality pipeline

You can use our *neurobattery* to reproduce our adolescent multiple modality analysis from the paper ^[http://www.nature.com/articles/sdata20153].  The neurobattery is online and administered by Jeffrey T. Duda ^[https://github.com/jeffduda/NeuroBattery].

# Brain mapping in the presence of lesions or other obstructions

Our fully automated registration and segmentation approach for the lesioned or occluded brain uses machine learning to identify and down-weight missing data.  One may also use pre-identified inclusive masks to focus registration on healthy tissue ^[` antsRegistration -d 3 ... etcetera ... -x healthyTissueMask.nii.gz` ].

# Statistical pipelines

*ANTs*-based statistical pipelines are best accessed through *ANTsR*.  We use these pipelines for:

* estimating cerebral blood flow with robust and data-driven methods

* performing structural and functional connectivity measurements

* employing biologically-regularized dimensionality reduction for hypothesis testing

* implementing exploratory multivariate analysis tools such as *sparse canonical correlation analysis for neuroimaging* ^[http://www.sciencedirect.com/science/article/pii/S1053811913009841]

* performing prediction studies ^[http://www.sciencedirect.com/science/article/pii/S1046202314003338]

* associating modalities ^[http://www.sciencedirect.com/science/article/pii/S105381191400891X]

and more basic studies such as mass univariate regression.

# Overview of *ANTs* programs

ANTs executables come in either binary ( compiled from `C++` ) or script form ( usually `bash` or `R` ).  Here, we summarize the relevant programs.


```{r, results='asis',echo=FALSE}
library(xtable)
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
############################################
prog=c("ANTS","see antsRegistration","old",
"ANTSAverage2DAffine.sh","see AverageAffine*","old",
"ANTSAverage3DAffine.sh","see AverageAffine*","old",
"ANTSIntegrateVectorField","ode integration of vector field","transformation",
"ANTSIntegrateVelocityField","ode integration of velocity field","transformation",
"ANTSJacobian","see CreateJacobianDeterminantImage","old",
"ANTSUseDeformationFieldToGetAffineTransform","fit affine tx to deformation field","old",
"ANTSUseLandmarkImagesToGetAffineTransform","fit affine tx to landmarks","transformation",
"ANTSUseLandmarkImagesToGetBSplineDisplacementField","fit bspline to landmarks","old",
"ANTSpexec.sh","helper for parallel execution","util",
"Atropos","EM segmentation framework and tools","segmentation",
"AverageAffineTransform","average affine transformations","transformation",
"AverageAffineTransformNoRigid","average affine transformations without rigid part","transformation",
"AverageImages","average list of images","processing",
"AverageTensorImages","average list of tensor images","processing",
"CheckTopology","see antsSurf","old",
"ClusterImageStatistics","summarize label images","old",
"ComposeMultiTransform","see antsApplyTransforms","old",
"CompositeTransformUtil","dis/assembles composite transform files","transformation",
"ConvertImage","change pixel type","old",
"ConvertImagePixelType","change pixel type","processing",
"ConvertInputImagePixelTypeToFloat","change pixel type","old",
"ConvertScalarImageToRGB","use lookup table to make RBG from gray","visualization",
"ConvertToJpg","make jpg from medical image","old",
"ConvertTransformFile","convert mat to txt, for example","transformation",
"CopyImageHeaderInformation","copy header from one image to another","util",
"CreateDTICohort","simulate DTI population","statistics",
"CreateDisplacementField","make vector field from component images","util",
"CreateImage","make an image","util",
"CreateJacobianDeterminantImage","compute deformation gradient from displacement","transformation",
"CreateTiledMosaic","tile images for viewing","visualization",
"CreateWarpedGridImage","make a warped grid from displacement","visualization",
"DeNrrd","?","?",
"DenoiseImage","non-local denoising","processing",
"ExtractRegionFromImage","get subset of image using indices","util",
"ExtractRegionFromImageByMask","get subset of image using mask","util",
"ExtractSliceFromImage","get slice from image","util",
"FitBSplineToPoints","fit bspline to point set","transformation",
"GetConnectedComponentsFeatureImages","?","processing",
"GetMeshAndTopology","see antsSurf","old",
"ITK_bld_internal_H5detect","?","util",
"ITK_bld_internal_H5make_libsettings","?","util",
"ImageCompare","see if images are nearly the same","util",
"ImageIntensityStatistics","simple summary stats for image +/- ROIs","statistics",
"ImageMath","basic processing operations on images","processing",
"ImageSetStatistics","compute mean, median, etc of images","statistics",
"KellyKapowski","image-based thickness estimator","processing",
"KellySlater","see KellyKapowski","old",
"LabelClustersUniquely","label each isolated region","segmentation",
"LabelGeometryMeasures","measure geometry of labeled regions","processing",
"LabelOverlapMeasures","compute overlaps","statistics",
"LaplacianThickness","old school thickness estimator","old",
"LesionFilling","WIP - do not use","old",
"MeasureImageSimilarity","similarity between image pairs","statistics",
"MeasureMinMaxMean","basic stats on an image","statistics",
"MemoryTest","memory profiler","util",
"MultiplyImages","multiply image1 by x","util",
"N3BiasFieldCorrection","ancient bias corrector","processing",
"N4BiasFieldCorrection","better bias corrector","processing",
"PasteImageIntoImage","put one image in another","util",
"PermuteFlipImageOrientationAxes","flip or permute image","util",
"PrintHeader","image header information","util",
"RebaseTensorImage","map tensor into new basis","processing",
"ReorientTensorImage","reorient tensors by transformation","transformation",
"ResampleImage","change image resolution","util",
"ResampleImageBySpacing","change image resolution","util",
"ResetDirection","change image direction to identity","util",
"SetDirectionByMatrix","change image direction","util",
"SetOrigin","set origin in image","util",
"SetSpacing","set spacing in image","util",
"SmoothImage","smooth in given units","processing",
"StackSlices","stack up a population of image slices","visualization",
"SurfaceBasedSmoothing","smoothing restricted to segmentation","processing",
"SurfaceCurvature","shape operator curvature","processing",
"TextureCooccurrenceFeatures","texture based statistics","processing",
"TextureRunLengthFeatures","texture based statistics","processing",
"ThresholdImage","simple image thresholding","segmentation",
"TileImages","collect images in tile form for viewing","visualization",
"TimeSCCAN","cca for temporal images","statistics",
"WarpImageMultiTransform","see antsApplyTransforms","old",
"WarpTensorImageMultiTransform","see antsApplyTransforms","old",
"WarpTimeSeriesImageMultiTransform","see antsApplyTransforms","old",
"WarpVTKPolyDataMultiTransform","see antsApplyTransformsToPoints","old",
"ants.sh","deprecated","old",
"antsAI","multi-start low-dimensional registration","transformation",
"antsASLProcessing.R","BMKandel ASL processing","processing",
"antsASLProcessing.sh","BMKandel ASL processing","processing",
"antsAffineInitializer","see antsAI","transformation",
"antsAlignOrigin","basic origin alignment","transformation",
"antsApplyTransforms","apply (multiple) transformations to image types","transformation",
"antsApplyTransformsToPoints","apply (multiple) transformations to points","transformation",
"antsAtroposN4.sh","joint segmentation and bias correction","segmentation",
"antsBOLDNetworkAnalysis.R","primive BOLD analysis - see ANTsR instead","old",
"antsBrainExtraction.sh","brain extraction via registration and segmentation","segmentation",
"antsCorticalThickness.sh","brain thickness pipeline","processing",
"antsIntermodalityIntrasubject.sh","example of how to map within subject","transformation",
"antsIntroduction.sh","see antsRegistrationSyNQuick.sh","old",
"antsJointFusion","better JLF implementation","segmentation",
"antsJointLabelFusion.sh","better JLF script","segmentation",
"antsLaplacianBoundaryCondition.R","WIP do not use","old",
"antsLongitudinalCorticalThickness.sh","longitudinal brain thickness pipeline","processing",
"antsMalfLabeling.sh","see antsJointLabelFusion","old",
"antsMotionCorr","time series motion correction","transformation",
"antsMotionCorrDiffusionDirection","DWI specific motion correction","transformation",
"antsMotionCorrExample","?","old",
"antsMotionCorrStats","summarize antsMotionCorr output","transformation",
"antsMultivariateTemplateConstruction.sh","multiple modality templates","transformation",
"antsMultivariateTemplateConstruction2.sh","multiple modality templates","transformation",
"antsNetworkAnalysis.R","basic network analysis - see ANTsR instead","old",
"antsNeuroimagingBattery","align MR modalities to common space","processing",
"antsRegistration","standard registration algorithms","transformation",
"antsRegistrationSpaceTime.sh","spatiotemporal registration methods","transformation",
"antsRegistrationSyN.sh","default decent quality registration","transformation",
"antsRegistrationSyNQuick.sh","default fast registration","transformation",
"antsSliceRegularizedRegistration","see spinal cord toolbox","transformation",
"antsSurf","surface rendering and other operations","visualization",
"antsTransformInfo","investigate a transformation","util",
"antsUtilitiesTesting","see how metric changes with tx","util",
"antsaffine.sh","see antsRegistrationSyN.sh","old",
"antsbashstats.sh","deprecated - see ANTsR","old",
"antsdeformationmag.sh","deprecated - see ANTsR","old",
"antsqsub.sh","deprecated","old",
"antswithdt.sh","deprecated","old",
"asymmetry.sh","basic asymmetry estimate - not recommended","old",
"basic_ants_example.sh","basic example - not recommended","old",
"bl","?","old",
"buildtemplateparallel.sh","old school but good template estimation","old",
"cbf_pasl_robust_batch.R","deprecated","old",
"cbf_pcasl_robust_batch.R","deprecated","old",
"compareTwoTransforms","as described","util",
"directlabels.sh","WIP - do not use","old",
"geodesicinterpolation.sh","shape-based interpolate two images","transformation",
"guidedregistration.sh","example of landmark based registration","transformation",
"iMath","utility","old",
"itkTestDriver","utility","old",
"jointfusion","deprecated","old",
"landmarkmatch.sh","example of landmark based registration","transformation",
"lohmann.sh","WIP - do not use","old",
"multi_template_script.sh","WIP - do not use","old",
"newAntsExample.sh","deprecated","old",
"optimalsmooth.sh","deprecated","old",
"phantomstudy.sh","deprecated - see ANTsR","old",
"registerimages.pl","deprecated","old",
"runprogramonimageset.pl","deprecated","old",
"sa","?","old",
"sccan","utility for sparse decomposition - see ANTsR","statistics",
"sccan_tests","WIP - do not use","old",
"shapeupdatetotemplate.sh","shape step for template","old",
"simpleSynRegistration","?","old",
"skel.sh","topological skeleton of segmentation","segmentation",
"submitexperimentalbuild.sh","?","old",
"sygnccavg.sh","deprecated","old",
"thickstudy.sh","deprecated","old",
"unbiased_longitudinal_map","deprecated","old",
"unbiased_pairwise_registration.sh","example of unbiased registration","old",
"unbiased_pairwise_registration_with_aux_images.sh","example of unbiased registration","old",
"waitForPBSQJobs.pl","utility","util",
"waitForSGEQJobs.pl","utility","util",
"waitForXGridJobs.pl","utility","util",
"warpimages.pl","deprecated","old",
"weightedaverage.pl","deprecated","old",
"weightedaverage.sh","deprecated","old",
"mrvnrf","machine learning segmentation - see ANTsR","segmentation")
nents=1:round(length(prog)/3)
mydata <- data.frame(
program=prog[  nents*3-2 ],
description=prog[  nents*3-1],
label=prog[nents*3]
  )
knitr::kable( mydata[ mydata$label == "transformation" , 1:2 ], 
              caption="ANTs transformation programs")
knitr::kable( mydata[ mydata$label == "segmentation" , 1:2 ], 
              caption="ANTs segmentation programs")
knitr::kable( mydata[ mydata$label == "processing" , 1:2 ], 
              caption="ANTs processing programs")
knitr::kable( mydata[ mydata$label == "statistics" , 1:2 ], 
              caption="ANTs statistics programs")
knitr::kable( mydata[ mydata$label == "visualization" , 1:2 ], 
              caption="ANTs visualization programs")
knitr::kable( mydata[ mydata$label == "util" , 1:2 ], 
              caption="ANTs utility programs")
# knitr::kable( mydata[21:40,] )
# knitr::kable( mydata[41:60,] )
```


[^ants]: http://stnava.github.io/ANTs/
[^antsr]: http://stnava.github.io/ANTsR/
[^tufte_latex]: https://code.google.com/p/tufte-latex/
[^books_be]: http://www.edwardtufte.com/tufte/books_be










