---
title: 'Introduction to *ANTsR*'
author: "Brian B. Avants et al."
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    colortheme: dolphin
    fonttheme: structurebold
    highlight: tango
    incremental: yes
    theme: AnnArbor
    includes:
      in_header: mystyle.tex
    toc: yes
---

```{r setup,eval=TRUE,results='hide',warning=FALSE,echo=FALSE}
# set this for your own compilation
bd="/Users/stnava/data/ANTsTutorial/"
options(width=50)
opts_chunk$set(fig.path='figure/antsr-',fig.align='center',fig.show='hold',size='footnotesize')
```


```{r setup0,results='hide',warning=FALSE,echo=FALSE,message=FALSE}
booleval<-FALSE
library(knitr)
library(rmarkdown)
options(width=50)
library( ANTsR )
library( oro.nifti )
library( visreg )
library( ggplot2 )
library( boot )
library( candisc )
library( pheatmap )
library( MASS )
ageregions<-"Not calculated"
myregion<-"CentralSulcus"
sccansig=0
```



# Intro

## ANTsR

* This slide or two briefly summarizes the [vignette](http://htmlpreview.github.io/?https://github.com/stnava/ANTsDoc/blob/master/html/ANTsR.html) for [*ANTsR*](http://stnava.github.io/ANTsR/)

* Interfaces state of the art image processing with R statistical methods.

* Wraps ANTs and ITK C++ core via Rcpp (Eddelbuettel 2013).

*  ITK is a templated C++ framework with I/O and support for arbitrary image types (usually 2, 3 or 4 dimensions) as well as surface representations.

* ITK/ANTs allow powerful image manipulations but lack a true statistical back-end.



## ANTsR Algorithms

In combination, these tools enable one to go from near-raw medical imaging data to a fully reproducible scientific publication (Avants 2015).

* core image processing and I/O: ITK (B. B. Avants, Tustison, et al. 2014);

* registration and utilities for image processing: ANTs mappings (Tustison, Cook, et al. 2014) and feature extraction (Tustison, Shrinidhi, et al. 2014);

*  dimensionality reduction: Eigenanatomy (Dhillon et al. 2014) and SCCAN (B. B. Avants, Libon, et al. 2014);

* methods for ASL-based cerebral blood flow quantification (Kandel et al. 2015);

* neighborhood representations of images that enable rich statistical models (Kandel et al. 2015)

* core statistics and temporal filtering via R packages that is amenable to BOLD image processing


## Setup

* Assume latest 3.x R-version from [CRAN](http://cran.r-project.org).

* Assume you have [*ANTsR*](https://github.com/stnava/ANTsR)

* `git clone http://github.com/stnava/RMI.git`

* `cd RMI/stnava`

* `git clone http://github.com/stnava/RMI_Data.git`

* in *R*, `install.packages("rmarkdown")`

* in *R*, `rmarkdown::render("rmi.Rmd")`

## Tutorial Goals
* Learn about *R*
in general (operations, dataframes, models) and *ANTsR*
* Understand basics of *R*
image-based statistics (I/O, accessing
values, structure, function)
* Practice some example reproducible studies ...
* Identify opportunities for innovation/future work
* A good source on data analysis with *R*
examples: [Advanced
Data Analysis from an Elementary Point of View](http://www.stat.cmu.edu/~cshalizi/uADA/12/lectures/ADAfaEPoV.pdf)



## The first 3 rules of statistics:

> Draw a picture, Draw a picture, Draw a picture.
> --Michael Starbird



## *R* in medical imaging?

Search "r-project.org $+$ medical $+$ imaging"

```{r rinmi}
dd<-read.csv( paste(bd,'data/RMI.csv',sep='') )
mdl<-lm( NPublications ~ Year +I(Year^2), data = dd)
```


## *R* in medical imaging?

*R* contains virtually all popular statistical and machine learning algorithms, including Boosting, the LASSO, and random
forests, often contributed by the inventors.
```{r vr,echo=FALSE,out.width='.75\\linewidth'}
visreg(mdl,main='Total Publications In Year X')
```



## Why use *R* in medical imaging?

* It's free --- not "free" like Matlab/SPM but really free

* It is the *de facto* standard for statistical computing

* a New York Times article from 2009 estimated
that there are at least 250,000 active R users

* Why not use Python?

* IMHO, *R* is easier to compile/maintain/install

* Visualization in *R* is as good or better than Python

* Most importantly---*statisticians contribute directly to R*

* Because of this, many Python users rely on *R*

* *R* facilitates reproducible research:

* [CRAN Task View](http://cran.r-project.org/web/views/ReproducibleResearch.html)

* [Biostatistics](http://biostatistics.oxfordjournals.org/content/10/3/405.long)

* [jstatsoft](http://www.jstatsoft.org/)

* *R* assists reproducible medical image analysis but it's still
effortful to do it correctly.  See
[comments on reproducibility in Frontiers](http://www.frontiersin.org/Journal/10.3389/fnins.2013.00162/full)



# Set up for Medical Image Analysis


##  We now open *R*

Install MIA-relevant packages.

```{r,echo=FALSE,results='hide'}
# some setup
options(width=60)  # make the printing fit on the page
set.seed(1121)   # make the results repeatable
library( ANTsR )
library( oro.nifti )
```


## Packages
```{r pkg}
pkgnames <- c( "visreg" , "boot" , "rgl",
 "knitr" , "ggplot2","oro.nifti" ,
 "candisc" , "pheatmap" )
# might try these too
pkgnames2=c( "abind", "BGLR", "caret", "cluster,
    d3Network", "DMwR", "e1071", "extremevalues", "fastICA",
    "fpc", "glasso", "glmnet", "grid", "igraph", "knitr",
    "magic", "MASS", "mFilter", "misc3d", "moments",
    "pixmap", "png", "psych", "randomForest", "rgl",
    "robust", "robustbase", "signal", "sna", "testthat",
    "visreg", "wmtsa")
k<-length( pkgnames )
```

## Next actually install the packages.

```bash
install.packages( pkgnames )
```


##

We installed `r k[1]` packages.


## What is `knitr` or `rmarkdown`?
* Yihui Xie's system
[http://yihui.name/knitr/](http://yihui.name/knitr/) for making
**documents that compute**

* `knitr` lets you write a document that employs *R*
directly

* *R*
evaluates code when the document is compiled

* The user controls when this does / does not happen ...

* Creates, figures, statistics etc that are embedded in rst, html,
latex, pretty much any common document format is doable.

## Simple `knitr` Example

Use `knitr` to make a couple of plots in our code "chunk":

```{r,boring-plots,fig.width=4,fig.height=4,out.width='.35\\linewidth',dev='pdf'}
x <- rnorm( 20 )
boxplot( x )
hist( x , main='', col="blue", probability = TRUE )
lines( density( x ), col = "red" )
```


## *R* $\star$Very$\star$ Basics

*R*
organizes data with dataframes, vectors, matrices and arrays (matrices with $\ge$
3 dimensions).

These can contain missing variables - but you must be careful about type!
```{r,out.width='.4\\linewidth'}
as.numeric( as.character( c("0.5",0.1,0.6,"A") ) )
```
A data frame is used for storing data tables. It is a list of vectors
of equal length.




## *R* $\star$Very$\star$ Basics 2

`mtcars` is a built-in *R*
dataframe
```{r,out.width='.4\\linewidth',results='show',echo=TRUE}
mtcars[c(1,13,28),1:6]
```

## MPG example

We analyze the relationship between MPG and other variables.

```{r,out.width='.4\\linewidth',results='hide'}
myform<-paste( colnames( mtcars )[2:ncol(mtcars)] ,
               collapse='+'  )
myform<-as.formula( paste( "mpg~", myform ) )
mdl <- lm(  myform , data = mtcars )
mdla<-stepAIC( mdl , direction =  c("both" ) )
print(myform)
```




## *R* $\star$Very$\star$ Basics 3

```{r,out.width='.4\\linewidth',results='show',echo=FALSE}
print(summary( lm(  mdla$call$formula , data = mtcars ) ) )
```
Vehicle weight ,  1/4 mile time  and Transmission type are most predictive!






## *R* $\star$Very$\star$ Basics: Draw a Picture

```{r,out.width='.3\\linewidth',dev='pdf',results='show',echo=TRUE}
mdl <- lm(  mdla$call$formula , data = mtcars )
visreg( mdl, xvar="wt")
visreg( mdl, xvar="qsec")
visreg( mdl, xvar="am")
```

**Oops! Transmission type should be a factor.**


## *R* $\star$Very$\star$ Basics: Draw a Picture Fix

```{r,out.width='.3\\linewidth',dev='pdf',results='show',echo=TRUE}
mtcars$am<-as.factor( mtcars$am )
mdl <- lm(  mdla$call$formula , data = mtcars )
visreg( mdl, xvar="wt")
visreg( mdl, xvar="qsec")
visreg( mdl, xvar="am")
```
This is better ...



## *R* $\star$Very$\star$ Basics: Draw a Picture 3

```{r,out.width='.6\\linewidth',dev='pdf',echo=TRUE,eval=TRUE}
 coplot(mpg ~ wt | qsec , data = mtcars, panel = panel.smooth, rows = 1)
```

**Exercise: Try this!**

# Medical Image Representations

## Data representation standards

Represent image as `antsImage`s which are basically ITK
type images with all I/O etc.
```{r antsimage}
img=antsImageRead( getANTsRData("r16" ) )
img
```

Can be converted to vectors, matrices, arrays ...

## Plot the `antsImage`

```{r plotai,echo=FALSE,out.width='.75\\linewidth'}
plot( img )
```

## Data representation standards

Represent an image as a *vector* (more on this later).  This
vector may be derived from a 2 or 3D array of spatially related voxels.
```{r,imgvec,out.width='.25\\linewidth',dev='pdf'}
imgmask=getMask( img )
mydat<-data.frame( space = 1:sum(imgmask==1),
                   imgvec = img[ imgmask == 1 ] )
ggplot(data=mydat, aes(x=space, y=imgvec, group=1)) + geom_line()
```



## Data Representation

Represent an image set as a *matrix*.

```{r imgmat}
fns=c( getANTsRData('r16'),
       getANTsRData('r27'),
       getANTsRData('r62') )
ilist=imageFileNames2ImageList( fns )
avg=antsAverageImages( ilist )
avgmask=getMask(avg)
imgmat=imageListToMatrix( ilist, avgmask )
print( dim(imgmat) )
```

## Show the average
```{r plotavg,echo=FALSE,out.width='.75\\linewidth'}
plot( avg )
```

## Matrix of vectorized images

```{r imgmat2,echo=FALSE,out.width='.75\\linewidth'}
pheatmap::pheatmap( (imgmat), cluster_rows=FALSE, cluster_cols=FALSE )
```


## Reading Images

Read em and weep

```{r,myio,echo=TRUE,warning=FALSE,results='show',fig.width=4,fig.height=4,out.width='.45\\linewidth'}
fn<-getANTsRData( "ch2" , usefixedlocation = FALSE  )

# oro.nifti
colin <- readNIfTI( fn )

# antsr
colina <- antsImageRead( fn )
```


## Quickly Show Images by oro.nifti
```{r,myviz,echo=TRUE,warning=FALSE,results='hide',fig.width=4,fig.height=4,out.width='.5\\linewidth'}
orthographic( as.array( colina ), oma=rep(2,4))
```

## Quickly Show Images w/ANTsR
```{r,myviza,echo=TRUE,warning=FALSE,tidy=TRUE,results='hide',fig.width=4,fig.height=4,out.width='.9\\linewidth'}
plot( as.antsImage( colin ) , axis=1 )
```



## Convert an Image to a Vector
Use *ANTsR* to convert an image to a vector.
```{r,convvec,echo=TRUE,warning=FALSE,tidy=TRUE,results='hide',fig.width=4,fig.height=4,out.width='.4\\linewidth'}
imgvec<-colina[  colina > 50  ]
print( length( imgvec ) )
```
Use  oro.nifti to convert an image to a vector.
```{r,convvec2,echo=TRUE,warning=FALSE,tidy=TRUE,results='hide',fig.width=4,fig.height=4,out.width='.4\\linewidth'}
imgvec<-colin[  colin > 50  ]
print( length( imgvec ) )
```
Both packages enable similar functionality in terms of accessing /
converting images to vectors.  *ANTsR* allows I/O to files other than
nifti such as meta, jpg, dicom, etc, anything ITK reads/writes.

# Quantifying Images


## Image Quantification with *R*

It is possible to implement full processing pipelines with *R*
for submission to distributed computing systems ...

My knowledge is limited to *ANTsR*.

There are also packages TractR, fMRI, dce-MRI and likely many more.

See the [medical imaging task view on CRAN](http://cran.r-project.org/web/views/MedicalImaging.html).


## The Basic Toolset from outside *R*

\textcolor{blue}{Registration: Data is in Examples/Data}
```bash
antsRegistration -d 2
 -m CC[r16slice.nii.gz,r64slice.nii.gz,1,4]
 -t SyN[0.25,3,0]  -o TEST -i 50x40x30 -c 50
```
`antsRegistration` in *ANTsR*

\textcolor{blue}{Segmentation}
```bash
Atropos -d 2 -a r16slice.nii.gz -x r16mask.nii.gz
-m [0.1,1x1]   -c [10,0]  -i kmeans[3]
-o [Output.nii.gz,Output\_prob\_\%02d.nii.gz]
```
`atropos` in *ANTsR*

`antsApplyTransforms` in both

## The Basic Toolset from outside *R*

\textcolor{blue}{Template building}
```bash
 bash buildtemplateparallel.sh -d 3 -m 30x50x20
-t GR  -s CC -c 1 -o OutPrefix  *ImageName*T1x.nii.gz
```
not available in *ANTsR*

# Vignettes

## ANTsR Vignette

[ANTsR vignette](http://htmlpreview.github.io/?https://github.com/stnava/ANTsDoc/blob/master/html/ANTsR.html)

## iMath Vignette

[iMath vignette](http://htmlpreview.github.io/?https://github.com/stnava/ANTsDoc/blob/master/html/iMath.html)

# The manual file

## Accessing help in ANTsR

```{r antsrhelp,eval=FALSE}
?sparseDecom
?antsRegistration
ANTsR::<double-tab>
```

Currently 199 help objects ...

Also see (little bit out of date) [manual](https://github.com/stnava/ANTsR/releases/download/v0.3.1/ANTsR-manual.pdf)
