---
title: 'Phantom Morphometry Study with *ANTsR*'
author: "Brian B. Avants et al."
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    colortheme: dolphin
    fonttheme: structurebold
    highlight: tango
    incremental: yes
    theme: AnnArbor
    includes:
      in_header: mystyle.tex
    toc: yes
  ioslides_presentation:
    highlight: tango
    incremental: yes
---

```{r setup,eval=TRUE,results='hide',warning=FALSE,echo=FALSE}
# set this for your own compilation
library(ANTsR)
bd="/Users/stnava/data/ANTsTutorial/"
templatefn=paste(bd,"phantomData/phantomtemplate.jpg",sep='')
if ( ! file.exists(templatefn) )
  stop( paste( "no image", templatefn ) )
allfns=Sys.glob( paste(bd,"phantomData/ph*wmgm.jpg",sep='')  )
demog=read.csv(paste(bd,'phantomData/phantpredictors.csv',sep=''))
ilist=imageFileNames2ImageList( allfns )
sccansig=0
```

# Phantom Morphometry Study

## Phantom Morphometry Study

* Manually drawn three tissue images

* Brain-like, two-dimensional slices

* Pre-built template (`./ANTsTutorial/phantomData/phantomtemplate.jpg`)

* We show how to do a log-jacobian based study

* ... and a tissue study.


## Look at population: S1

```{r pop,results='asis'}
plot( ilist[[1]], window.img=c(0,255) )
```

## Look at population: S2

```{r pop2,results='asis'}
plot( ilist[[2]], window.img=c(0,255) )
```

## Look at population: S5

```{r pop5,results='asis'}
plot( ilist[[5]], window.img=c(0,255) )
```

## Look at population: S6

```{r pop6,results='asis'}
plot( ilist[[6]], window.img=c(0,255) )
```

# Segmentation

## First: segment the images

```{r segment}
template=antsImageRead( templatefn )
ilist=imageFileNames2ImageList( allfns )
plist=imageFileNames2ImageList( allfns )
for ( x in 1:length(ilist) )
  {
  mask=getMask( ilist[[x]] )  %>% iMath("MD",1)
  img = ilist[[ x ]]
  km=kmeansSegmentation(  img, 3, mask ) # segment
  gm=km$probabilityimages[[2]]
  plist[[x]]=gm
  }
```

## Quick look at segmentation results

```{r segment1,echo=FALSE}
plot( ilist[[1]])
```

## Quick look at segmentation results

```{r segment1b,echo=FALSE}
mask=getMask( ilist[[1]] )  %>% iMath("MD",1)
img = ilist[[ 1 ]]
km=kmeansSegmentation(  img, 3, mask ) # segment
plot( km$segmentation, window.img=c(0,3) )
```


# Registration

## Second: register the images

```{r register}
template=antsImageRead( templatefn )
jlist=imageFileNames2ImageList( allfns )
for ( x in 1:length(ilist) )
  {
  gm=plist[[x]]
  tx=antsRegistration( template, ilist[[x]], "SyNCC" )
  gmw=antsApplyTransforms( template, gm, tx$fwdtransforms)
  plist[[x]]=gmw
  jlist[[x]]=createJacobianDeterminantImage( template,
    tx$fwdtransforms[1], 1) %>% smoothImage(1.5)
  }
```

## Quick look at the template ...

```{r templ,echo=FALSE}
plot( template )
```


## And a registration

```{r regtempl,echo=FALSE}
plot( tx$warpedmovout  )
```

## And a registration overlay

```{r regtemplb,echo=FALSE}
plot( template, tx$warpedmovout %>% iMath("Canny",1,5,12),
  window.overlay=c(0.5,1)  )
```

# Jacobian


##  Now look at a log-jacobian image ...

```{r logj,echo=FALSE}
plot( jlist[[1]])
```

##  another log-jacobian image ...

```{r logjb,echo=FALSE}
plot( jlist[[5]])
```


## Now we can do statistics on either warped segmentations or the log-jacobian

.... but we should look at the demographics first.



# Statistics

## Always the first thing ...

Look at the demographics file ...

```{r demog,echo=FALSE}
knitr::kable(demog)
```

## Let's do a quick test on the jacobian

Check out `bigLMStats`. It is a great little function for
large (on the left) statistical studies ...

```{r jaclm}
templatemask=getMask( template )
jmat=imageListToMatrix( jlist, templatemask )
mdl=lm( jmat ~ vol + gender, data=demog )
mdl=bigLMStats( mdl )
volstats=mdl$beta.pval[ "vol", ]
qv=p.adjust(volstats,'BH')
qvimg=makeImage( templatemask, 1-qv )
```

for large on the right studies, use a quick
implementation in `RcppEigen` or `RcppArmadillo`

## Look at results

```{r jaclmviz,echo=FALSE}
plot( template, qvimg, window.overlay=c(0.95,1))
```


## Exercise ... repeat for the GM ...

...

```{r gmlm,echo=FALSE}
gmat=imageListToMatrix( plist, templatemask )
mdl=lm( gmat ~ vol + gender, data=demog )
mdl=bigLMStats( mdl )
gmstats=mdl$beta.pval[ "vol", ]
gqv=p.adjust(gmstats,'BH')
gqvimg=makeImage( templatemask, 1-gqv )
```


## Exercise ... use a global covariate ...

... this is always important to think about ...


# Multivariate statistics

## Try a multivariate version of the same study

Do we believe that many aspects of cognition are fundamentally multivariate?


##  .... multivariate statistics in *R*

```{r,multivar,echo=TRUE,eval=TRUE,results='hide',warning=FALSE}
mypreds<-scale( as.matrix( cbind( demog$vol*(-1) ) ) )
scaledMatrix = scale( jmat )
scaledMatrix = scaledMatrix - min( scaledMatrix)
sccan<-sparseDecom2(
  inmatrix  = list( scaledMatrix, mypreds ),
  inmask    = c( templatemask , NA ) ,
  mycoption = 1,
  sparseness=c( 0.2, -1 ),
  nvecs=1, its=10,
  cthresh=c(50,0),
  smooth=0.0, perms=200 )
knitr::kable(sccan$ccasummary)
sccansol <- sccan$eig1[[1]] %>% iMath("Normalize")
```
Exercise: What happens to the permutation based significance
when you vary parameters?

##  Visualizing multivariate statistics in *R*

```{r sccanviz,echo=FALSE,out.width='.6\\linewidth',message=FALSE,warning=FALSE,eval=TRUE}
plot( template, sccansol, window.overlay=c(0.1,1.01) )
```

# And MNI space ....

## Mapping to MNI space with *ANTsR*

* We implemented a simple approach to mapping a
custom template, and its blobs, to MNI coordinates.

* See `?getTemplateCoordinates` for an example.

* We also report summary regional labels.

* First, get a template, here the `ch2` brain.
```{r ch2}
if ( ! exists("tem") )
  tem<-antsImageRead( getANTsRData("ch2") )
```

## Mapping to MNI space with *ANTsR*

* Next, define pseudoclusters in ch2 space.
```{r defineclust}
clust <- thresholdImage( tem, 82, 90 ) %>%
   iMath("ME",1)  %>% labelClusters()
```

## Mapping to MNI space with *ANTsR*

* Download the MNI template and its labels.  
```{r mymnidl,eval=TRUE}
if ( ! exists("mymni") ) {
# try getANTsRData if you have www access
  mymni<-list( antsImageRead(getANTsRData("mni") ),
               antsImageRead(getANTsRData("mnib") ),
               antsImageRead(getANTsRData("mnia") ) )
}
```

## Mapping to MNI space with *ANTsR*

* Next define a pair of your template and its clusters.

```{r mypair,eval=TRUE}
template_cluster_pair<-list(tem,clust)
```

* Call the function to map to MNI space.
```{r mymni,eval=TRUE}
t1=Sys.time()
gcoords<-getTemplateCoordinates( template_cluster_pair ,
    mymni , convertToTal = TRUE )
t2=Sys.time()
print(t2-t1)
```

* This performs a registration and returns the coordinates for each cluster.


## Reporting cluster coordinates in MNI space with *ANTsR*

\small

Show the table of results.
```{r repomni,echo=FALSE,eval=TRUE}
knitr::kable( gcoords$templatepoints  )
```

## Review

* A simple reproducible example for a phantom morphometry study

* Plotting, segmentation, registration

* Statistics of two flavors ....

* Mapping to MNI coordinates.

