---
title: 'Phantom Morphometry Study with *ANTsR*'
author: "Brian B. Avants et al."
date: "May 7, 2015"
output:
  beamer_presentation:
    colortheme: dolphin
    fonttheme: structurebold
    highlight: tango
    incremental: yes
    theme: AnnArbor
    includes:
      in_header: mystyle.tex
    toc: yes
  ioslides_presentation:
    highlight: tango
    incremental: yes
---

```{r setup,eval=TRUE,results='hide',warning=FALSE,echo=FALSE}
# set this for your own compilation
library(ANTsR)
bd="/Users/stnava/data/ANTsTutorial/"
templatefn=paste(bd,"phantomData/phantomtemplate.jpg",sep='')
if ( ! file.exists(templatefn) )
  stop( paste( "no image", templatefn ) )
allfns=Sys.glob( paste(bd,"phantomData/ph*wmgm.jpg",sep='')  )
demog=read.csv(paste(bd,'phantomData/phantpredictors.csv',sep=''))
ilist=imageFileNames2ImageList( allfns )
sccansig=0
```

## Phantom Morphometry Study

* Manually drawn three tissue images

* Brain-like, two-dimensional slices

* Pre-built template (`./ANTsTutorial/phantomData/phantomtemplate.jpg`)

* We show how to do a log-jacobian based study

* ... and a tissue study.

## Look at population: S1

```{r pop,results='asis'}
plot( ilist[[1]], window.img=c(0,255) )
```

## Look at population: S2

```{r pop2,results='asis'}
plot( ilist[[2]], window.img=c(0,255) )
```

## Look at population: S5

```{r pop5,results='asis'}
plot( ilist[[5]], window.img=c(0,255) )
```

## Look at population: S6

```{r pop6,results='asis'}
plot( ilist[[6]], window.img=c(0,255) )
```

## First: segment the images

```{r segment}
template=antsImageRead( templatefn )
ilist=imageFileNames2ImageList( allfns )
plist=imageFileNames2ImageList( allfns )
for ( x in 1:length(ilist) )
  {
  mask=getMask( ilist[[x]] )  %>% iMath("MD",1)
  img = ilist[[ x ]]
  km=kmeansSegmentation(  img, 3, mask ) # segment
  gm=km$probabilityimages[[2]]
  plist[[x]]=gm
  }
```
## Quick look at segmentation results

```{r segment1,echo=FALSE}
plot( ilist[[1]])
```

## Quick look at segmentation results

```{r segment1b,echo=FALSE}
mask=getMask( ilist[[1]] )  %>% iMath("MD",1)
img = ilist[[ 1 ]]
km=kmeansSegmentation(  img, 3, mask ) # segment
plot( km$segmentation, window.img=c(0,3) )
```

## Second: register the images

```{r register}
template=antsImageRead( templatefn )
jlist=imageFileNames2ImageList( allfns )
for ( x in 1:length(ilist) )
  {
  gm=plist[[x]]
  tx=antsRegistration( template, ilist[[x]], "SyNCC" )
  gmw=antsApplyTransforms( template, gm, tx$fwdtransforms)
  plist[[x]]=gmw
  jlist[[x]]=createJacobianDeterminantImage( template,
    tx$fwdtransforms[1], 1) %>% smoothImage(1.5)
  }
```

## Quick look at the template ...

```{r templ,echo=FALSE}
plot( template )
```


## And a registration

```{r regtempl,echo=FALSE}
plot( tx$warpedmovout  )
```

## And a registration overlay

```{r regtemplb,echo=FALSE}
plot( template, tx$warpedmovout %>% iMath("Canny"),
  window.overlay=c(0.5,1)  )
```

##  Now look at a log-jacobian image ...

```{r logj,echo=FALSE}
plot( jlist[[1]])
```

##  another log-jacobian image ...

```{r logjb,echo=FALSE}
plot( jlist[[5]])
```

## Now we can do statistics on either warped segmentations or the log-jacobian

Look at the demographics file ...

```{r demog,echo=FALSE}
knitr::kable(demog)
```

## Let's do a quick test on the jacobian

`bigLMStats` is a great little function for
large (on the left) statistical studies ...

```{r jaclm}
templatemask=getMask( template )
jmat=imageListToMatrix( jlist, templatemask )
mdl=lm( jmat ~ vol + gender, data=demog )
mdl=bigLMStats( mdl )
volstats=mdl$beta.pval[ "vol", ]
qv=p.adjust(volstats,'BH')
qvimg=makeImage( templatemask, 1-qv )
```

for large on the right studies, use a quick
implementation in `RcppEigen` or `RcppArmadillo`

## Look at results

```{r jaclmviz,echo=FALSE}
plot( template, qvimg, window.overlay=c(0.99,1))
```

## Exercise ... repeat for the GM ...

...

```{r gmlm,echo=FALSE}
gmat=imageListToMatrix( plist, templatemask )
mdl=lm( gmat ~ vol + gender, data=demog )
mdl=bigLMStats( mdl )
gmstats=mdl$beta.pval[ "vol", ]
gqv=p.adjust(gmstats,'BH')
gqvimg=makeImage( templatemask, 1-gqv )
```


## Exercise ... use a global covariate ...

... this is always important to think about ...

## Try a multivariate version of the same study


##  .... multivariate statistics in *R*
\transdissolve
```{r,multivar,echo=TRUE,eval=TRUE,results='hide',warning=FALSE}
mypreds<-as.matrix( cbind( demog$vol*(-1), demog$vol*(-1) ) )
sccan<-sparseDecom2(
  inmatrix  = list( jmat, mypreds ),
  inmask    = c( templatemask , NA ) ,
  mycoption = 1,
  sparseness=c( 0.2, -1 ),
  nvecs=1, its=10,
  cthresh=c(10,0),
  smooth=0.0, perms=200 )
knitr::kable(sccan$ccasummary)
sccansol <- sccan$eig1[[1]] %>% iMath("Normalize")
```
Exercise: What happens to the permutation based significance
when you vary parameters?

##  Visualizing multivariate statistics in *R*
\transdissolve

```{r sccanviz,echo=FALSE,out.width='.6\\linewidth',message=FALSE,warning=FALSE,eval=TRUE}
plot( template, sccansol, window.overlay=c(0.1,1.01) )
```
