---
title: 'The Pediatric Template of Brain Perfusion: Quick reproducible analysis'
author: "Brian B. Avants et al."
date: "May 7, 2015"
output:
  beamer_presentation:
    colortheme: dolphin
    fonttheme: structurebold
    highlight: zenburn
    incremental: yes
    theme: AnnArbor
    includes:
      in_header: mystyle.tex
    toc: yes
---

# Overview

## Overview 
The Pediatric Template of Brain Perfusion (PTBP) [at figshare](http://figshare.com/articles/The_Pediatric_Template_of_Brain_Perfusion_PTBP_/923555).

- Free multiple modality MRI data with demographics and psychometrics
- The data is accompanied by an [organized csv file](http://files.figshare.com/1699436/ptbp_summary_demographics.csv) 
- The full data is available at [figshare](http://figshare.com/articles/The_Pediatric_Template_of_Brain_Perfusion_PTBP_/923555)
- Here we use a slightly processed version of the data
- Each image contains a slab version of T1, FA, B0, CBF, Thickness, Segmentation and AAL in subject space
- There is also a template contained in the download.


## Download the processed data
\Huge

- Download
- <https://www.dropbox.com/s/u8bg5z5k6do56om/tutorial_data_pad.zip?dl=0>
- Then unzip the file.

# Quick Look 

## If you have not already, download [*ANTsR*](http://stnava.github.io/ANTsR/)

* Get *R* from [OSX *R*](http://cran.r-project.org/bin/macosx/) or [Linux *R*](http://www.jason-french.com/blog/2013/03/11/installing-r-in-linux/)
* Install [*RStudio*](http://www.rstudio.com/products/rstudio/download/)
* Download the *ANTsR* tar.gz to `filename.tar.gz`
* Install via: 
```bash
   R CMD INSTALL filename.tar.gz
```
* Test via ( in `R`) calling: 
```r
  library(ANTsR)
```
* Done!

## Slide with R Code and Output

```{r}
summary(cars)
```

## Slide with Plot

```{r, echo=FALSE}
plot(cars)
```

