---
title: 'The Pediatric Template of Brain Perfusion: Quick introduction'
author: "Brian B. Avants et al."
date: "May 7, 2015"
output:
  beamer_presentation:
    colortheme: dolphin
    fonttheme: structurebold
    highlight: tango
    incremental: yes
    theme: AnnArbor
    includes:
      in_header: mystyle.tex
    toc: yes
--- 

# Overview and resources

## Overview 

The Pediatric Template of Brain Perfusion (PTBP) [at figshare](http://figshare.com/articles/The_Pediatric_Template_of_Brain_Perfusion_PTBP_/923555).

- Free multiple modality MRI data with demographics and psychometrics
- Collected to aid as a reference for normal variability in structure and cerebral blood flow during adolescence
- The data is accompanied by an [organized csv file](http://files.figshare.com/1699436/ptbp_summary_demographics.csv) 
- The full data is available at [figshare](http://figshare.com/articles/The_Pediatric_Template_of_Brain_Perfusion_PTBP_/923555)
- Here we use a slightly processed version of the data
- Each subject contains a slab version of T1, FA, B0, CBF, Thickness, Segmentation and AAL in subject space
- There is also a template contained in the download.

## Download the processed data

\Large

- Download
- <https://www.dropbox.com/s/u8bg5z5k6do56om/tutorial_data_pad.zip?dl=0>
- Then unzip the file.

## If you have not already, download [*ANTsR*](http://stnava.github.io/ANTsR/)

* Get *R* from [OSX *R*](http://cran.r-project.org/bin/macosx/) or 
  [Linux *R*](http://www.jason-french.com/blog/2013/03/11/installing-r-in-linux/), 
  optionally [*RStudio*](http://www.rstudio.com/products/rstudio/download/)
* Install *ANTsR* dependencies
```r
pkgmin=c("magrittr","Rcpp") # core dependencies
pkgxtra=c("igraph","randomForest","misc3d","rgl",
  "mFilter","fastICA","fpc","knitr","rmarkdown",
  "pixmap","png","signal","visreg")
install.packages( c( pkgmin, pkgxtra ) ) 
```
* Download the *ANTsR* tar.gz to `filename.tar.gz`
* Install via command line (or *RStudio* `tools` $\rightarrow$ `install`): 
```bash
   R CMD INSTALL filename.tar.gz
```
* Test via ( in `R` ) calling: 
```r
  library(ANTsR)
```
* Done!

# Quick Look 

## Get the demographics file

```{r csv}
library(ANTsR)
bd="/Users/stnava/data/ANTsTutorial/"
csvlink="http://files.figshare.com/1699436/ptbp_summary_demographics.csv"
tfn=tempfile(fileext='.csv')
demog = read.csv( paste( bd, "data/ptbp_summary_demographics.csv", sep='' ) )
if ( ! exists("demog") )
  {
  demog=download.file(csvlink,tfn)
  if ( demog == 0 ) demog=read.csv( tfn )
  }
```

## Investigate the demographics

```{r demog}
str(demog[,1:10])
```

## Investigate the demographics

```{r demog2}
str(demog[,11:20])
```

# Review multiple modality images

## Read the structural PTBP template image

```{r readstruct, echo=TRUE}
t1fn = paste(bd,
  "./data/template/PTBP_T1_BrainCerebellum.nii.gz",
  sep='')
t1tem = antsImageRead( t1fn )
```

## Plot the structural PTBP template image

```{r plotstruct, echo=FALSE, warning=FALSE}
plot(t1tem,dorot=1,slices=8)
```

## Plot the CBF PTBP template image

```{r plotcbf, echo=FALSE, warning=FALSE}
cbftem = paste(bd,"./data/template/PTBP_CBF.nii.gz",sep='')
cbftem = antsImageRead( cbftem )
plot(cbftem,dorot=1,slices=8)
```

## Find subject image

```{r substruct, echo=TRUE}
subnum=11
id=demog$SubID[subnum]
dt=demog$ScanDate[subnum]
exts=c( "mprage_t1.nii.gz" , 
        "fa_anatomical.nii.gz", 
        "MeanCBFWarpedToT1.nii.gz", 
        "CorticalThickness.nii.gz", 
        "BrainSegmentation.nii.gz", 
        "AAL.nii.gz" )
pre=paste( bd,"data/Subjects/",id,"/",dt,"/*",sep='')
fns=Sys.glob( paste( pre, exts , sep='') )
```

## Review each modality
```{r mod1, echo=FALSE}
plot( antsImageRead( fns[1] ), dorot=1, slices=8)
```

## Review each modality 
```{r mod2, echo=FALSE}
plot( antsImageRead( fns[2] ), dorot=1, slices=8)
```

## Review each modality 
```{r mod3, echo=FALSE}
plot( antsImageRead( fns[3] ), dorot=1, slices=8)
```

## Review each modality 
```{r mod4, echo=FALSE}
plot( antsImageRead( fns[4] ), dorot=1, slices=8)
```

## Review each modality 
```{r mod5, echo=FALSE}
plot( antsImageRead( fns[5] ), dorot=1, slices=8)
```

## Review each modality 
```{r mod5b, echo=FALSE}
cseg=antsImageRead( fns[5] ) %>% thresholdImage(2,2)
plot( antsImageRead( fns[1] ), cseg, dorot=1, slices=8)
```

## Review each modality 
```{r mod6, echo=FALSE}
aal=antsImageRead( fns[6] )
plot( antsImageRead( fns[1] ), aal, dorot=1, slices=8, 
      window.overlay=c( 1, max(aal) ) )
```

## Review each modality 
```{r mod1b, echo=FALSE}
fa=antsImageRead( fns[2] ) %>% iMath("Canny")
plot( antsImageRead( fns[1] ), fa, dorot=1, slices=8 )
```

## Count complete subject images

```{r ctsubimages, echo=TRUE}
havecompletesubject=rep(FALSE,nrow(demog))
for ( x in 1:nrow(demog) )
  {
  id = demog$SubID[x]
  dt = demog$ScanDate[x]
  pre = paste( bd,"data/Subjects/",id,"/",dt,"/*",sep='')
  fns = Sys.glob( paste( pre, exts , sep='') )
  if ( length( fns ) == 6 ) havecompletesubject[x]=TRUE
  }
nsub=sum( havecompletesubject  )
```

## PTBP in our tutorial

We have `r nsub` multiple modality subjects to process
and to analyze.

We will use them to:

* Build a template

* Construct template priors with [joint label fusion](http://www.ncbi.nlm.nih.gov/pubmed/22732662)

* Normalize, segment, compute thickness

* Map other modalities to structural reference

* Investigate [Eigenanatomy](https://scholar.google.com/scholar?hl=en&q=Eigenanatomy&btnG=&as_sdt=1%2C5&as_sdtp=) and [SCCAN](https://scholar.google.com/scholar?q=sparse+canonical+correlation+avants&btnG=&hl=en&as_sdt=0%2C5) for relating demographic and psychometric measurements to the imaging and ...

* relating imaging modalities to each other.

* Finally, put all this together to create a reproducible analysis document for the PTBP.



## Template priors with SyN-Quick

```{r maptem, echo=TRUE, eval=FALSE, warning=FALSE}
# t1tem read before
ilist=list()
seglist=list()
for ( x in which( havecompletesubject )[1:60] )
  {
  id = demog$SubID[x]
  dt = demog$ScanDate[x]
  pre = paste( bd,"data/Subjects/",id,"/",dt,"/*",sep='')
  fns = Sys.glob( paste( pre, exts , sep='') )
  t1 = antsImageRead( fns[1] )
  mytx = antsRegistration( t1tem, t1, "SyN" )
  seg = antsImageRead( fns[5] )
  segw = antsApplyTransforms( t1tem, seg, mytx$fwd, interpolator="NearestNeighbor"  )
  ilist = lappend( ilist, mytx$warpedmovout )
  seglist = lappend( seglist, segw )
  }
```



## Template [Joint Label/Intensity Fusion](http://journal.frontiersin.org/article/10.3389/fninf.2013.00027/abstract)

```{r jlftem, echo=TRUE, eval=FALSE, warning=FALSE }
mk=getMask(t1tem)
jlf=jointIntensityFusion3D( t1tem, mk, ilist, 
  labelList=seglist, rad=rep(2,3), rSearch=2, 
  computeProbs=TRUE, includezero=TRUE)
# write out results
tf=tempfile()
print(tf)
antsImageWrite( jlf$segimg, paste(tf,'seg.nii.gz',sep='') )
antsImageWrite( jlf$predimg, paste(tf,'pred.nii.gz',sep='') )
for ( i in 1:length(jlf$probimgs) )
  antsImageWrite( jlf$probimgs[[i]], 
                  paste(tf,'prob',i,'.nii.gz',sep='') )
```



