---
title: 'The Pediatric Template of Brain Perfusion: MRVNRF with *ANTsR*'
author: "Brian B. Avants et al."
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    colortheme: dolphin
    fonttheme: structurebold
    highlight: tango
    incremental: yes
    theme: AnnArbor
    includes:
      in_header: mystyle.tex
    toc: yes
  ioslides_presentation:
    highlight: tango
    incremental: yes
---

```{r setup,eval=TRUE,results='hide',warning=FALSE,echo=FALSE}
# set this for your own compilation
set.seed(1000)
bd="/Users/stnava/data/ANTsTutorial/"
```

# MRVNRF with ANTsR

## Extension of a recent paper

[**BRATS Challenge**](http://www.ncbi.nlm.nih.gov/pubmed/25433513)

> Tustison et al

* difference 1: here, using multiple resolution RFs

* difference 2: predicting a continuous outcome ...


## Basic Setup

```{r dataio}
library(ANTsR)
library(visreg)
library(randomForest)
thkmask=antsImageRead( paste(bd,"data/ptbp_mask_thickness.nii.gz",sep='') )
famask=antsImageRead( paste(bd,"data/ptbp_mask_fa.nii.gz",sep='') )
demog=read.csv("/Users/stnava/data/ANTsTutorial/data/ptbp_summary_demographics.csv")
demog=demog[ , 1:19 ]
```

## Get all voxel data

```{r form}
thkmat=antsImageRead( paste(bd,"data/ptbp_vox_thk.mha",sep='') )
famat=antsImageRead( paste(bd,"data/ptbp_vox_fa.mha",sep='') )
cbfmat=antsImageRead( paste(bd,"data/ptbp_vox_cbf.mha",sep='') )
thkmat=as.matrix( thkmat )
cbfmat=as.matrix( cbfmat )
famat=as.matrix( famat )
```

## Get common images and make a dataframe

```{r mdf}
wp=( !is.na(rowMeans(thkmat)) & rowMeans(thkmat) > 0.5  &
     !is.na(rowMeans(famat)) & rowMeans(famat) > 0.2  &
     !is.na(rowMeans(cbfmat)) & rowMeans(cbfmat) > 40 )
mydf=data.frame( AgeAtScan=demog[wp,]$AgeAtScan,
                 PIQ=demog[wp,]$Performance.IQ,
                 VIQ=demog[wp,]$Verbal.IQ,
                 BV=demog[wp,]$BV,
                 Cortex=demog[wp,]$Cortex,
                 LadderCom=demog[wp,]$Teen.Ladder.Community.Score,
                 LadderSES=demog[wp,]$Teen.Ladder.SES.score,
                 Income=demog[wp,]$Income )
for ( kk in 2:ncol(mydf) ) mydf[,kk]=antsrimpute( mydf[,kk] )
```  


## Train/Test

First get the data organized.
```{r orgpred}
library(randomForest)
groups <- rep( c(1,2), 1000 )[1:nrow(mydf)]
traing=groups==1
testg=groups==2
```

## Select parts of masks that are related to age ( our outcome )


```{r fsel}
temp=thkmat[wp,]
thkcor=abs( cor(temp[traing,], mydf$VIQ[traing]) )
temp=famat[wp,]
facor=abs( cor(temp[traing,], mydf$VIQ[traing]) )
temp=cbfmat[wp,]
cbfcor=abs( cor(temp[traing,], mydf$VIQ[traing]) )
corimg=makeImage( thkmask, thkcor ) %>% smoothImage(1)
corimg=corimg+makeImage( thkmask, cbfcor ) %>% smoothImage(1)
corimg=corimg+makeImage( famask, facor ) %>% smoothImage(1)
mrvmask=thresholdImage( corimg, 0.25, Inf )
rm(temp)
```

## MRVNRF Mask

```{r mrviz3,echo=FALSE}
plot( thkmask, corimg, axis=3, nslices=16, window.overlay=c(0.2,0.66) )
```


## Set up mrvnrfs

```{r mrvsubs}
slist=list()
agelist=list()
ct=1
for ( x in which(wp) )
  {
  if (  wp[x] == TRUE )
    {
    i1=makeImage(thkmask, thkmat[x,])
    i2=makeImage(famask, famat[x,])
    i3=makeImage(thkmask, cbfmat[x,] )
    feati=list( i1 )
    slist[[ct]]=feati
    agelist[[ct]]=mydf$VIQ[x]
    ct=ct+1
    }
  }
rm( thkmat )
rm( famat )
rm( cbfmat )
gc()
```

## MRVNRF Train

We train from the CBF, FA and cortical thickness images.

```{r mrvtrain}
y=mydf$VIQ[traing]
x=slist[traing]
mr=c(4,2)
if ( ! exists("rfm") )
rfm<-mrvnrfs( y, x,
  mrvmask, rad=rep(1,3),
  nsamples=200, ntrees=2000,
  multiResSchedule=mr,
  asFactors = FALSE )
```


## MRVNRF Test

```{r mrvtest}
if ( ! exists("rfm2") )
rfm2<-mrvnrfs.predict( rfm$rflist,
  slist[testg], mrvmask, rad=rep(1,3),
  multiResSchedule=mr,
  asFactors = FALSE )
cor( mydf$VIQ[testg] , rfm2$seg )
print(  mean( abs( mydf$VIQ[testg] - rfm2$seg) ) )
```

## MRVNRF Viz

```{r mrviz}
predmat<-imageListToMatrix( unlist(rfm2$probs) , mrvmask )
rfpred <- apply( predmat,  FUN = median, MARGIN = 1)
print(  cor.test( mydf$VIQ[testg] , rfpred ) )
print(  mean( abs( mydf$VIQ[testg] - rfpred) ) )
corrmat<-antsrimpute( cor( predmat , mydf$VIQ[testg] ) )
corimg<-makeImage( mrvmask, corrmat )
```

## MRVNRF Viz

```{r mrviz2,echo=FALSE}
plot( thkmask, corimg, axis=3, nslices=16,
   window.overlay=c( 0.1 , max(corimg) ) )
```
