---
title: 'The Pediatric Template of Brain Perfusion: Eigenanatomy with *ANTsR*'
author: "Brian B. Avants et al."
date: "May 7, 2015"
output:
  beamer_presentation:
    colortheme: dolphin
    fonttheme: structurebold
    highlight: tango
    incremental: yes
    theme: AnnArbor
    includes:
      in_header: mystyle.tex
    toc: yes
  ioslides_presentation:
    highlight: tango
    incremental: yes
---

```{r setup,eval=TRUE,results='hide',warning=FALSE,echo=FALSE}
# set this for your own compilation
bd="/Users/stnava/data/ANTsTutorial/"
```

# Eigenanatomy with ANTsR

## Based on a recent paper

[**Eigenanatomy: Sparse dimensionality reduction for multi-modal medical image analysis**](https://scholar.google.com/citations?view_op=view_citation&hl=en&user=a2sMoDEAAAAJ&citation_for_view=a2sMoDEAAAAJ:qjMakFHDy7sC)

> BM Kandel, DJJ Wang, JC Gee, BB Avants*, "Methods", 2015

* difference 1: here, using some rank reduction to speed things up.

* difference 2: ben's work uses some eigenvector fusion to reduce
the impact of parameter choice ...

* we might do a little of that below.

## Basic Setup

```{r dataio}
library(ANTsR)
library(visreg)
neanatvecs=20
thkmask=antsImageRead( paste(bd,"data/ptbp_mask_thickness.nii.gz",sep='') )
qth=0.05
mth='BH'
demog=read.csv("/Users/stnava/data/ANTsTutorial/data/ptbp_summary_demographics.csv")
demog=demog[ , 1:19 ]
```

## Set up the formula

```{r form}
formulathkbase=" AgeAtScan * Sex + I(AgeAtScan^2)"
formulathkbase=" PIQ + BV"
formulathktest=" AgeAtScan * Sex + I(AgeAtScan^2)"
voi="AgeAtScan"
```

## Generic version: Function

```{r genericvox}
voxandeanatstudy <- function( demog, imgmat, imgmask,
                              formulabase, formulatest,
                              voi,
                              exclusionThresh,
                              baddata,
                              outprefix,
                              nv )
  {
  imgrowmeans=rowMeans( imgmat )
  wp = rep( FALSE, nrow( demog ) )
  for ( sub in unique( demog$SubID ) ) # get unique subjects
    {
    ww=which( demog$SubID == sub )
    ww=ww[ !is.na( imgrowmeans[ww] )  &
             imgrowmeans[ww] > exclusionThresh ]
    if ( length( ww ) > 0 ) wp[ ww[ 1 ] ] = TRUE
    }
#  if ( ! all( is.na(baddata) ) ) # FIXME
#   wp[  which(wp==TRUE)[ baddata ] ]=FALSE # bad data
  ilist=matrixToImages( imgmat[wp,], imgmask )
#  for ( i in 1:length(ilist) ) ilist[[i]]=smoothImage( ilist[[i]], 2.0 )
  mydf=data.frame(
                 Sex=demog[wp,]$Sex,
                 AgeAtScan=demog[wp,]$AgeAtScan,
                 PIQ=demog[wp,]$Performance.IQ,
                 VIQ=demog[wp,]$Verbal.IQ,
                 BV=demog[wp,]$BV,
                 Cortex=demog[wp,]$Cortex,
                 LadderCom=demog[wp,]$Teen.Ladder.Community.Score,
                 LadderSES=demog[wp,]$Teen.Ladder.SES.score,
                 Income=demog[wp,]$Income )
  for ( kk in 2:ncol(mydf) ) mydf[,kk]=antsrimpute( mydf[,kk] )
  imgmat=imageListToMatrix( ilist, imgmask )
  locform=formula( paste( "imgmat ~", formulabase ,"+", formulatest ) )
  mdl=lm( locform, data=mydf )
  voxlm=bigLMStats( mdl , 1.e-8 )
  print(paste("begin low",outprefix) )
  lowmat = lowrankRowMatrix( imgmat , 10 )
  print(paste("begin eanat",outprefix) )
  imgeanat=sparseDecom( lowmat, imgmask, nvecs=nv, smooth=0,
    sparseness=0.05, cthresh=2000, mycoption=0, its=5 )
  rm( lowmat )
  print(paste("end eanat",outprefix) )
  eproj=abs( imageListToMatrix( imgeanat$eig , imgmask )  )
  eproj=( eproj/rowSums(eproj) )
  imgproj = imgmat %*% t( eproj  )
  mydf=data.frame( mydf, imgproj )
  print( names(mydf) )
  formbase=formula( paste( "imgproj ~", formulabase ) )
  formtest=formula( paste( "imgproj ~", formulabase ,"+", formulatest ) )
  mdl1=lm( formbase, data=mydf )
  mdl2=lm( formtest, data=mydf )
  eanatlm=bigLMStats( mdl2 , 1.e-6 )
  eseg=eigSeg( mask=imgmask, imgeanat$eig, F  )
  ofn=paste( outprefix, '_eseg.nii.gz', sep='' )
  antsImageWrite( eseg , ofn )
  anv=anova( mdl1, mdl2)
  return( list(mydf=mydf,
               voxlm=voxlm,
               eanatlm=eanatlm,
               eseg=eseg,
               imgproj=imgproj,
               whichSubjects=wp,
               eanat=imgeanat,
               anv=anv) )
  }
```  


# Three little modalities


## Thickness Eanat

```{r thkfun}
############################################
thkmask=antsImageRead( paste(bd,"data/ptbp_mask_thickness.nii.gz",sep='') )
thkfn=paste(bd,"data/ptbp_vox_thk.mha",sep='')
thkmat=as.matrix( antsImageRead( thkfn ) )
baddata=NA
ethk=voxandeanatstudy( demog, thkmat, thkmask,
       formulathkbase, formulathktest, voi=voi,
       exclusionThresh = 0.5, baddata=baddata,
       outprefix='/tmp/ASS_THK', nv=neanatvecs )
qv=p.adjust( ethk$eanatlm$beta.pval[ voi, ], method='BH' )
print( min(qv) )
```

## Thickness Eanat: Anova
```{r thkfunanv}
print( ethk$anv )
```

## FA Eanat

```{r fafun}
fafn=paste(bd,"data/ptbp_vox_fa.mha",sep='')
famask=antsImageRead( paste(bd,"data/ptbp_mask_fa.nii.gz",sep='') )
famat=as.matrix( antsImageRead( fafn ) )
baddata=c(42,44) # FA
efa=voxandeanatstudy( demog, famat, famask,
       formulathkbase, formulathktest, voi=voi,
       exclusionThresh = 0.2, baddata=baddata,
       outprefix='/tmp/ASS_FA', nv=neanatvecs )
qv=p.adjust( efa$eanatlm$beta.pval[ voi, ], method='BH' )
print( min(qv) )
```

## FA Eanat: Anova
```{r fafunanv}
print( efa$anv )
```

## CBF Eanat

```{r cbffun}
cbffn=paste(bd,"data/ptbp_vox_cbf.mha",sep='')
cbfmat=as.matrix( antsImageRead( cbffn ) )
baddata=NA
ecbf=voxandeanatstudy( demog, cbfmat, thkmask,
       formulathkbase, formulathktest, voi=voi,
       exclusionThresh = 45, baddata=baddata,
       outprefix='/tmp/ASS_CBF', nv=neanatvecs )
qv=p.adjust( ecbf$eanatlm$beta.pval[ voi,], method='BH' )
print( min(qv) )
```

## CBF Eanat: Anova
```{r cbffunanv}
print( ecbf$anv )
```

# Three little modalities jointly

## Make a new function that recomputes projections from existing Eanat

```{r eanatproj}
eanatproj <- function( demog, imgmat,
  imgmask, imgeanat,
  formulabase, formulatest,
  voi, whichSubs )
  {
  wp = whichSubs
  mydf=data.frame(
                 Sex=demog[wp,]$Sex,
                 AgeAtScan=demog[wp,]$AgeAtScan,
                 PIQ=demog[wp,]$Performance.IQ,
                 VIQ=demog[wp,]$Verbal.IQ,
                 BV=demog[wp,]$BV,
                 Cortex=demog[wp,]$Cortex,
                 LadderCom=demog[wp,]$Teen.Ladder.Community.Score,
                 LadderSES=demog[wp,]$Teen.Ladder.SES.score,
                 Income=demog[wp,]$Income )
  for ( kk in 2:ncol(mydf) ) mydf[,kk]=antsrimpute( mydf[,kk] )
  imgmats=imgmat[wp,]
  locform=formula( paste( "imgmats ~", formulabase ,"+", formulatest ) )
  eproj=abs( imageListToMatrix( imgeanat$eig , imgmask )  )
  eproj=( eproj/rowSums(eproj) )
  imgproj = imgmats %*% t( eproj  )
  mydf=data.frame( mydf, imgproj )
  formbase=formula( paste( "imgproj ~", formulabase ) )
  formtest=formula( paste( "imgproj ~", formulabase ,"+", formulatest ) )
  mdl1=lm( formbase, data=mydf )
  mdl2=lm( formtest, data=mydf )
  eanatlm=bigLMStats( mdl2 , 1.e-6 )
  print( anova( mdl1, mdl2) )
  return( mydf )
  }
```


## Combined: Option 1 $=$ Merge

Here, we have to identify which subjects have all modalities.

```{r modmerge}
wth=ethk$whichSubjects
wfa=efa$whichSubjects
wcb=ecbf$whichSubjects
haveAll=( wth & wfa & wcb )
```

## Call the new function ...


```{r eprojcall}
thproj=eanatproj( demog, thkmat, thkmask,
  imgeanat=ethk$eanat,
  formulathkbase, formulathktest,
  voi=voi, whichSubs=haveAll )
cbproj=eanatproj( demog, cbfmat, thkmask,
  imgeanat=ecbf$eanat,
  formulathkbase, formulathktest,
  voi=voi, whichSubs=haveAll )
faproj=eanatproj( demog, famat, famask,
  imgeanat=efa$eanat,
  formulathkbase, formulathktest,
  voi=voi, whichSubs=haveAll )
```

## Join them all

```{r joiner}
n=ncol(faproj)
faprojsub=faproj[,(n-neanatvecs+1):n]
n=ncol(cbproj)
cbprojsub=cbproj[,(n-neanatvecs+1):n]
alldf=cbind( thproj, FA=faprojsub, CBF=cbprojsub)
```


## Joint stats: Thickness, FA $+$ CBF

```{r jointstats}
mdl1=lm( AgeAtScan ~ . ,
  data=alldf[,c(1:2,3,5,8)])
mdl2=lm( AgeAtScan ~ . ,
  data=alldf[,c(1:2,3,5,8,10:ncol(alldf))])
anova(mdl1,mdl2)
mdl2=lm( AgeAtScan ~ . ,
  data=alldf[,c(1:2,3,5,8,10:(10+neanatvecs-1))])
anova(mdl1,mdl2)
```


## Do some `visreg` ?

...

## Combined: Option 2 $=$ Impute

Did not implement this well.... but relatively easy to improve ...

```{r modimpute}
wth=ethk$whichSubjects
wfa=efa$whichSubjects
wcb=ecbf$whichSubjects
haveAll=( wth | wfa | wcb )
```


## Combined: Option 2 $=$ Impute

Did not implement this well.... but relatively easy to improve ...

```{r doimp}
thkmati=thkmat
thkmati[ haveAll, ]=antsrimpute( thkmat[ haveAll, ] )
famati=famat
famati[ haveAll, ]=antsrimpute( famat[ haveAll, ] )
cbfmati=cbfmat
cbfmati[ haveAll, ]=antsrimpute( cbfmat[ haveAll, ] )
```



## Call the new function ... imputation

```{r eprojcalli}
thproj=eanatproj( demog, thkmati, thkmask,
  imgeanat=ethk$eanat,
  formulathkbase, formulathktest,
  voi=voi, whichSubs=haveAll )
cbproj=eanatproj( demog, cbfmati, thkmask,
  imgeanat=ecbf$eanat,
  formulathkbase, formulathktest,
  voi=voi, whichSubs=haveAll )
faproj=eanatproj( demog, famati, famask,
  imgeanat=efa$eanat,
  formulathkbase, formulathktest,
  voi=voi, whichSubs=haveAll )
```

## Join them all ... imputation

```{r joineri}
n=ncol(faproj)
faprojsub=faproj[,(n-neanatvecs+1):n]
n=ncol(cbproj)
cbprojsub=cbproj[,(n-neanatvecs+1):n]
alldf=cbind( thproj, FA=faprojsub, CBF=cbprojsub)
```


## Joint stats ... imputation

```{r jointstatsi}
mdl1=lm( AgeAtScan ~ . ,
  data=alldf[,c(1:2,3,5,8)])
mdl2=lm( AgeAtScan ~ . ,
  data=alldf[,c(1:2,3,5,8,10:ncol(alldf))])
anova(mdl1,mdl2)
mdl2=lm( AgeAtScan ~ . ,
  data=alldf[,c(1:2,3,5,8,10:(10+neanatvecs-1))])
anova(mdl1,mdl2)
```

## PIQ stats?

```{r piq,eval=FALSE,echo=FALSE}
mdl1=lm( PIQ ~ . , data=alldf[,c(1:2,3,5,8)])
mdl2=lm( PIQ ~ . , data=alldf[,c(1:2,3,5,8,10:ncol(alldf))])
anova(mdl1,mdl2)
mdl2=lm( PIQ ~ . , data=alldf[,c(1:2,3,5,8,10:(10+neanatvecs-1))])
anova(mdl1,mdl2)
```

# Review

## Discussion

* ...

* ...

* ...
