---
title: 'The Pediatric Template of Brain Perfusion: Eigenanatomy with *ANTsR*'
author: "Brian B. Avants et al."
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    colortheme: dolphin
    fonttheme: structurebold
    highlight: tango
    incremental: yes
    theme: AnnArbor
    includes:
      in_header: mystyle.tex
    toc: yes
  ioslides_presentation:
    highlight: tango
    incremental: yes
---

```{r setup,eval=TRUE,results='hide',warning=FALSE,echo=FALSE}
# set this for your own compilation
bd="/Users/stnava/data/ANTsTutorial/"
set.seed(11)
library( candisc )
library(ggplot2)
```

# Classical multivariate methods updated for neuroimaging


## What is multiple regression?
The solution to a quadratic minimization problem:

**Multiple Regression**
$$
\|  y - X \beta \|^2  +  \textcolor{blue}{\lambda \| \beta \| ^2}
$$
Solved by ordinary least squares methods:
$$
\hat{\beta}=( X^T X)^{-1} X^T y
$$
with theory for turning $\beta$ entries into "significance" measurements.

The "ridge" penalty is useful if $p >> n$.



## Principal Component Analysis
Also the solution to a quadratic minimization problem:

**PCA**: $U$, $V$ minimize reconstruction error:
$$
\|  X  - U V^T \|^2   +  \textcolor{blue}{\sum_k \lambda_k \| V_k
\|_1}
$$
Each of the columns of $X$ is a linear combination of the
columns of $U$.

## Principal Component Analysis

Easy solution in *R*
(w/o penalties):
```bash
mysolution=svd(X)
# or prcomp(X) if X not centered
```

The $\ell_1$ penalty is useful if $p >> n$.



## Principal Component Analysis Example
```{r,out.width='.8\\linewidth',eval=TRUE,show='hide',echo=FALSE}
PCbiplot <- function(PC, x="PC1", y="PC2") {
    # PC being a prcomp object
    data <- data.frame(obsnames=row.names(PC$x), PC$x)
    plot <- ggplot(data, aes_string(x=x, y=y)) + geom_text(alpha=.4, size=3, aes(label=obsnames))
    plot <- plot + geom_hline(aes(0), size=.2) + geom_vline(aes(0), size=.2)
    datapc <- data.frame(varnames=rownames(PC$rotation), PC$rotation)
    mult <- min(
        (max(data[,y]) - min(data[,y])/(max(datapc[,y])-min(datapc[,y]))),
        (max(data[,x]) - min(data[,x])/(max(datapc[,x])-min(datapc[,x])))
        )
    datapc <- transform(datapc,
            v1 = .7 * mult * (get(x)),
            v2 = .7 * mult * (get(y))
            )
    plot <- plot + coord_equal() + geom_text(data=datapc, aes(x=v1, y=v2, label=varnames), size = 5, vjust=1, color="red")
    plot <- plot + geom_segment(data=datapc, aes(x=0, y=0, xend=v1, yend=v2), alpha=0.75, color="red")
    plot
}
fit <- prcomp(USArrests, scale=T)
```

```{r,out.width='.85\\linewidth',dev='pdf',echo=FALSE,eval=TRUE}
PCbiplot(fit)
```




## Canonical Correlation Analysis

**CCA** Generalizes Multiple Regression

$$  
Y V + \textcolor{blue}{\sum_k \lambda_k \| V_k \|_1} \propto X W  + \textcolor{blue}{\sum_k \gamma_k \| W_k \|_1}
$$
where $Y, V, X, W$ are matrices and $V, W$ are canonical variates (the
CCA solutions).


## Canonical Correlation Analysis
Also easy in *R*
(SVD used internally):
```{r,out.width='.4\\linewidth',echo=TRUE,eval=TRUE,results='hide'}
 enginedata<-mtcars[,c(2,3,4,11)]
 outputdata<-mtcars[,c(1,7)]
 mycca<-cancor( enginedata, outputdata )
```

CCA is "symmetric" in that the sets X and Y have equivalent
status.  A truly multivariate multiple regression.



## Canonical Correlation Analysis Visualization 1
```{r,out.width='.7\\linewidth',dev='pdf',echo=TRUE,eval=TRUE,warning=FALSE,results='show'}
print( mycca )
```




## Canonical Correlation Analysis Visualization 2
```{r,out.width='.65\\linewidth',dev='pdf',echo=TRUE,eval=TRUE,warning=FALSE,results='hide'}
heplot(mycca, xpd=TRUE, scale=0.3)
```


## Sparse multiple regression/PCA/CCA
Any of the methods can be made sparse by enforcing the penalties
previously highlighted in \textcolor{blue}{blue}.

**Sparse Optimization**

* Formulate the problem as a constrained optimization.
* Identify the gradient descent solution---*without sparseness*.
* Use projected gradient descent to solve the optimization---*with sparseness*.
* In imaging, other constraints are valuable too.



## *R* Multivariate Study with `SCCAN`

* Sparse canonical correlation analysis for neuroimaging

* Compare training/testing reliability wrt e.g. voxel-wise univariate feature selection

* Problem: find  voxels in training that predict X in testing

* A separate tutorial is here [sccan tutorial](http://stnava.github.io/sccanTutorial/) that uses the same data.



# Eigenanatomy with ANTsR

## Based on a recent paper

[**Eigenanatomy: Sparse dimensionality reduction for multi-modal medical image analysis**](https://scholar.google.com/citations?view_op=view_citation&hl=en&user=a2sMoDEAAAAJ&citation_for_view=a2sMoDEAAAAJ:qjMakFHDy7sC)

> BM Kandel, DJJ Wang, JC Gee, BB Avants*, "Methods", 2015

* difference 1: here, using some rank reduction to speed things up.

* difference 2: ben's work uses some eigenvector fusion to reduce
the impact of parameter choice ...

* we might do a little of that below.

## Basic Setup

```{r dataio}
library(ANTsR)
library(visreg)
neanatvecs=2
thkmask=antsImageRead( paste(bd,"data/ptbp_mask_thickness.nii.gz",sep='') )
qth=0.05
mth='BH'
demog=read.csv(paste(bd,"data/ptbp_summary_demographics.csv",sep='') )
demog=demog[ , 1:19 ]
```

## Set up the formula

```{r form}
formulathkbase=" AgeAtScan * Sex + I(AgeAtScan^2)"
formulathkbase=" PIQ + BV"
formulathktest=" AgeAtScan * Sex + I(AgeAtScan^2)"
voi="AgeAtScan"
```

## Generic version: Function

```{r genericvox}
voxandeanatstudy <- function( demog, imgmat, imgmask,
                              formulabase, formulatest,
                              voi,
                              exclusionThresh,
                              baddata,
                              outprefix,
                              nv )
  {
  imgrowmeans=rowMeans( imgmat )
  wp = rep( FALSE, nrow( demog ) )
  for ( sub in unique( demog$SubID ) ) # get unique subjects
    {
    ww=which( demog$SubID == sub )
    ww=ww[ !is.na( imgrowmeans[ww] )  &
             imgrowmeans[ww] > exclusionThresh ]
    if ( length( ww ) > 0 ) wp[ ww[ 1 ] ] = TRUE
    }
#  if ( ! all( is.na(baddata) ) ) # FIXME
#   wp[  which(wp==TRUE)[ baddata ] ]=FALSE # bad data
  ilist=matrixToImages( imgmat[wp,], imgmask )
#  for ( i in 1:length(ilist) ) ilist[[i]]=smoothImage( ilist[[i]], 2.0 )
  mydf=data.frame(
                 Sex=demog[wp,]$Sex,
                 AgeAtScan=demog[wp,]$AgeAtScan,
                 PIQ=demog[wp,]$Performance.IQ,
                 VIQ=demog[wp,]$Verbal.IQ,
                 BV=demog[wp,]$BV,
                 Cortex=demog[wp,]$Cortex,
                 LadderCom=demog[wp,]$Teen.Ladder.Community.Score,
                 LadderSES=demog[wp,]$Teen.Ladder.SES.score,
                 Income=demog[wp,]$Income )
  for ( kk in 2:ncol(mydf) ) mydf[,kk]=antsrimpute( mydf[,kk] )
  imgmat=imageListToMatrix( ilist, imgmask )
  locform=formula( paste( "imgmat ~", formulabase ,"+", formulatest ) )
  mdl=lm( locform, data=mydf )
  voxlm=bigLMStats( mdl , 1.e-8 )
  print(paste("begin low",outprefix) )
  lowmat = lowrankRowMatrix( imgmat , 10 )
  print(paste("begin eanat",outprefix) )
  imgeanat=sparseDecom( lowmat, imgmask, nvecs=nv, smooth=0.5,
    sparseness=0.05, cthresh=20000, mycoption=0, its=5 )
  rm( lowmat )
  print(paste("end eanat",outprefix) )
  eproj=abs( imageListToMatrix( imgeanat$eig , imgmask )  )
  eproj=( eproj/rowSums(eproj) )
  imgproj = imgmat %*% t( eproj  )
  mydf=data.frame( mydf, imgproj )
  print( names(mydf) )
  formbase=formula( paste( "imgproj ~", formulabase ) )
  formtest=formula( paste( "imgproj ~", formulabase ,"+", formulatest ) )
  mdl1=lm( formbase, data=mydf )
  mdl2=lm( formtest, data=mydf )
  eanatlm=bigLMStats( mdl2 , 1.e-6 )
  eseg=eigSeg( mask=imgmask, imgeanat$eig, F  )
  ofn=paste( outprefix, '_eseg.nii.gz', sep='' )
  antsImageWrite( eseg , ofn )
  anv=anova( mdl1, mdl2)
  return( list(mydf=mydf,
               voxlm=voxlm,
               eanatlm=eanatlm,
               eseg=eseg,
               imgproj=imgproj,
               whichSubjects=wp,
               eanat=imgeanat,
               anv=anv) )
  }
```  


# Three little modalities


## Thickness Eanat

```{r thkfun}
############################################
thkmask=antsImageRead( paste(bd,"data/ptbp_mask_thickness.nii.gz",sep='') )
thkfn=paste(bd,"data/ptbp_vox_thk.mha",sep='')
thkmat=as.matrix( antsImageRead( thkfn ) )
baddata=NA
if ( ! exists("ethk") )
ethk=voxandeanatstudy( demog, thkmat, thkmask,
       formulathkbase, formulathktest, voi=voi,
       exclusionThresh = 0.5, baddata=baddata,
       outprefix='/tmp/Eanat_THK', nv=neanatvecs )
qv=p.adjust( ethk$eanatlm$beta.pval[ voi, ], method='BH' )
print( min(qv) )
```

## Thickness Eanat: Anova
```{r thkfunanv}
knitr::kable( ethk$anv )
```

## FA Eanat

```{r fafun}
fafn=paste(bd,"data/ptbp_vox_fa.mha",sep='')
famask=antsImageRead( paste(bd,"data/ptbp_mask_fa.nii.gz",sep='') )
famat=as.matrix( antsImageRead( fafn ) )
baddata=c(42,44) # FA
if ( ! exists("efa") )
efa=voxandeanatstudy( demog, famat, famask,
       formulathkbase, formulathktest, voi=voi,
       exclusionThresh = 0.2, baddata=baddata,
       outprefix='/tmp/Eanat_FA', nv=neanatvecs )
qv=p.adjust( efa$eanatlm$beta.pval[ voi, ], method='BH' )
print( min(qv) )
```

## FA Eanat: Anova
```{r fafunanv}
knitr::kable( efa$anv )
```

## CBF Eanat

```{r cbffun}
cbffn=paste(bd,"data/ptbp_vox_cbf.mha",sep='')
cbfmat=as.matrix( antsImageRead( cbffn ) )
baddata=NA
if ( ! exists("ecbf") )
ecbf=voxandeanatstudy( demog, cbfmat, thkmask,
       formulathkbase, formulathktest, voi=voi,
       exclusionThresh = 45, baddata=baddata,
       outprefix='/tmp/Eanat_CBF', nv=neanatvecs )
qv=p.adjust( ecbf$eanatlm$beta.pval[ voi,], method='BH' )
print( min(qv) )
```

## CBF Eanat: Anova
```{r cbffunanv}
knitr::kable( ecbf$anv )
```

# Visualize eigenanatomy segmentations

## Thickness Viz
```{r thkv}
plot( thkmask, ethk$eseg , axis=3 )
```


## CBF Viz
```{r cbfv}
plot( thkmask, ecbf$eseg , axis=3 )
```


## FA Viz
```{r fav}
plot( famask, efa$eseg , axis=3 )
```


# Three little modalities jointly

## Make a new function that recomputes projections from existing Eanat

```{r eanatproj}
eanatproj <- function( demog, imgmat,
  imgmask, imgeanat,
  formulabase, formulatest,
  voi, whichSubs )
  {
  wp = whichSubs
  mydf=data.frame(
                 Sex=demog[wp,]$Sex,
                 AgeAtScan=demog[wp,]$AgeAtScan,
                 PIQ=demog[wp,]$Performance.IQ,
                 VIQ=demog[wp,]$Verbal.IQ,
                 BV=demog[wp,]$BV,
                 Cortex=demog[wp,]$Cortex,
                 LadderCom=demog[wp,]$Teen.Ladder.Community.Score,
                 LadderSES=demog[wp,]$Teen.Ladder.SES.score,
                 Income=demog[wp,]$Income )
  for ( kk in 2:ncol(mydf) ) mydf[,kk]=antsrimpute( mydf[,kk] )
  imgmats=imgmat[wp,]
  locform=formula( paste( "imgmats ~", formulabase ,"+", formulatest ) )
  eproj=abs( imageListToMatrix( imgeanat$eig , imgmask )  )
  eproj=( eproj/rowSums(eproj) )
  imgproj = imgmats %*% t( eproj  )
  mydf=data.frame( mydf, imgproj )
  formbase=formula( paste( "imgproj ~", formulabase ) )
  formtest=formula( paste( "imgproj ~", formulabase ,"+", formulatest ) )
  mdl1=lm( formbase, data=mydf )
  mdl2=lm( formtest, data=mydf )
  eanatlm=bigLMStats( mdl2 , 1.e-6 )
  knitr::kable( anova( mdl1, mdl2) )
  return( mydf )
  }
```


## Modalities Combined: Option 1 $=$ Merge

Here, we have to identify which subjects have all modalities.

```{r modmerge1}
wth=ethk$whichSubjects
wfa=efa$whichSubjects
wcb=ecbf$whichSubjects
haveAllMerge=( wth & wfa & wcb )
```

## Call the new function ...


```{r eprojcall}
thproj=eanatproj( demog, thkmat, thkmask,
  imgeanat=ethk$eanat,
  formulathkbase, formulathktest,
  voi=voi, whichSubs=haveAllMerge )
cbproj=eanatproj( demog, cbfmat, thkmask,
  imgeanat=ecbf$eanat,
  formulathkbase, formulathktest,
  voi=voi, whichSubs=haveAllMerge )
faproj=eanatproj( demog, famat, famask,
  imgeanat=efa$eanat,
  formulathkbase, formulathktest,
  voi=voi, whichSubs=haveAllMerge )
```

## Join them all

```{r joiner}
n=ncol(faproj)
faprojsub=faproj[,(n-neanatvecs+1):n]
n=ncol(cbproj)
cbprojsub=cbproj[,(n-neanatvecs+1):n]
alldf=cbind( thproj, FA=faprojsub, CBF=cbprojsub)
```


## Joint stats: Thickness, FA $+$ CBF

```{r jointstats,echo=FALSE}
mdl1=lm( AgeAtScan ~ . ,
  data=alldf[,c(1:2,3,5,8)])
mdl2=lm( AgeAtScan ~ . ,
  data=alldf[,c(1:2,3,5,8,10:ncol(alldf))])
knitr::kable(anova(mdl1,mdl2))
```


## Do some `visreg` ?

...

## Let's do some prediction

Use random forests to decide how relevant each
modality is to reproducible predictions of age.

First get the data organized.
```{r orgpred}
library(randomForest)
```



## Build the RF model


```{r rfpred,echo=FALSE}
groups <- rep( c(1,2), 1000 )[1:nrow(alldf)]
traing=groups==1
testg=groups==2
mdl=randomForest(  AgeAtScan ~ . , data=alldf[traing,], ntrees=2000 )
predage = predict( mdl, newdata=alldf[ testg, ] )
print( mean( abs(predage - alldf[testg,]$AgeAtScan ) ) )
temp=data.frame(predage=predage,realage=alldf[testg,]$AgeAtScan )
vmdl=lm( predage~realage,data=temp)
visreg::visreg(vmdl)
```

## RF importance plot

```{r rfimp,echo=FALSE}
impdf=data.frame(
  predNames=colnames(alldf)[ rev(order(mdl$importance)) ],
  importance=mdl$importance[ rev(order(mdl$importance)) ]
  )
knitr::kable(  impdf )
```


## Build the RF model: Just thickness

```{r rfpredt,echo=FALSE}
mdl=randomForest(  AgeAtScan ~ . , data=thproj[traing,] )
predage = predict( mdl, newdata=thproj[ testg, ] )
print( mean( abs(predage - thproj[testg,]$AgeAtScan ) ) )
temp=data.frame(predage=predage,realage=thproj[testg,]$AgeAtScan )
vmdl=lm( predage~realage,data=temp)
visreg::visreg(vmdl)
```


## Modalities Combined: Option 2 $=$ Impute

Here, we have to identify which subjects have all modalities.

```{r modmerge}
wth=ethk$whichSubjects
wfa=efa$whichSubjects
wcb=ecbf$whichSubjects
haveAll=( wth | wfa | wcb )
```

## Do the imputation

Did not implement this well.... but relatively easy to improve ...

```{r doimp}
thkmati=thkmat
thkmati[ haveAll, ]=antsrimpute( thkmat[ haveAll, ] )
famati=famat
famati[ haveAll, ]=antsrimpute( famat[ haveAll, ] )
cbfmati=cbfmat
cbfmati[ haveAll, ]=antsrimpute( cbfmat[ haveAll, ] )
```


## Call the new function ... imputation

```{r eprojcalli}
thproj=eanatproj( demog, thkmati, thkmask,
  imgeanat=ethk$eanat,
  formulathkbase, formulathktest,
  voi=voi, whichSubs=haveAll )
cbproj=eanatproj( demog, cbfmati, thkmask,
  imgeanat=ecbf$eanat,
  formulathkbase, formulathktest,
  voi=voi, whichSubs=haveAll )
faproj=eanatproj( demog, famati, famask,
  imgeanat=efa$eanat,
  formulathkbase, formulathktest,
  voi=voi, whichSubs=haveAll )
```

## Join them all ... imputation

```{r joineri}
n=ncol(faproj)
faprojsub=faproj[,(n-neanatvecs+1):n]
n=ncol(cbproj)
cbprojsub=cbproj[,(n-neanatvecs+1):n]
alldf=cbind( thproj, FA=faprojsub, CBF=cbprojsub)
```


## Joint stats ... imputation

```{r jointstatsi}
mdl1=lm( AgeAtScan ~ . ,
  data=alldf[,c(1:2,3,5,8)])
mdl2=lm( AgeAtScan ~ . ,
  data=alldf[,c(1:2,3,5,8,10:ncol(alldf))])
knitr::kable( anova(mdl1,mdl2) )
```


# SCCAN with ANTsR

## A little bit of CCA ...

First, thickness ...

```{r thkcca,eval=TRUE}
sdemog=antsrimpute( demog[haveAllMerge,c(7,8)] )
# above is a bug sparseDecom/2 does not handle NAs well
temp=thkmat[haveAllMerge,][traing,]
inmatrix=list( temp ,
  data.matrix(sdemog[traing,])  )
if ( ! exists("sccan") )
  sccan=sparseDecom2(inmatrix, c(thkmask, NA),
     sparseness = c( 0.005, 0.3 ),
     nvecs = 2, its = 15, cthresh = c(5000, 0),
     perms = 0, smooth = 0.5, mycoption = 1 )
```

## Check the CCA regions

```{r thkcca4,eval=TRUE,echo=FALSE}
ccaeseg=eigSeg( thkmask, sccan$eig1, FALSE  )
plot( thkmask, ccaeseg, axis=3, nslices=16 )
```

## Look over cca results

Which demographic variables match which imaging variables?

```{r thkcca2,eval=TRUE}
rownames( sccan$eig2 )=colnames( sdemog )
knitr::kable( sccan$eig2 )
```

## Quick CCA-based prediction

PIQ prediction ...

```{r thkcca3,eval=TRUE,echo=FALSE}
sccanmat=imageListToMatrix( sccan$eig1, thkmask)
sccanmat=sccanmat/rowSums(sccanmat)
sccanproj = thkmat[haveAllMerge,] %*% t( sccanmat )
ccadf=data.frame( sdemog, sccanproj,
  gender=demog$Sex[haveAllMerge],
  age=demog$AgeAtScan[haveAllMerge])
mdl=randomForest( Performance.IQ ~ . ,
  data=ccadf[traing,c(2,3:6)] )
pviq = predict( mdl,
  newdata=ccadf[ testg,c(2,3:6)] )
print( mean( abs( pviq - ccadf[testg,]$Performance.IQ) ) )
cor.test( pviq, ccadf[testg,]$Performance.IQ )
```


## More cca ...

Second, FA and CBF ...

These modalities relate strongly to age ... how do they covary?

```{r faccacbf,eval=TRUE}
sdemog=antsrimpute( demog[haveAllMerge,c(3,7,8)] )
temp=famat[haveAllMerge,][traing,]
for ( i in 1:nrow(temp) )
  temp[i,]=temp[i,]-mean(temp[i,])
temp2=cbfmat[haveAllMerge,][traing,]
for ( i in 1:nrow(temp2) )
    temp2[i,]=temp2[i,]-mean(temp2[i,])
inmatrix=list( temp , temp2 )
if ( ! exists("fcsccan") )
  fcsccan=sparseDecom2(inmatrix, c(famask, thkmask),
     sparseness = c( 0.05, -0.05 ),
     nvecs = 3, its = 2, cthresh = c(5000, 5000),
     perms = 0, smooth = 0.5, mycoption = 1 )
```


## Check the CCA regions:FA

```{r faccavfa,eval=TRUE,echo=FALSE}
ccaeseg=eigSeg( famask, fcsccan$eig1, FALSE  )
plot( famask, ccaeseg, axis=3, nslices=16 )
```

## Check the CCA regions:CBF

```{r faccavcbf,eval=TRUE,echo=FALSE}
ccaeseg=eigSeg( thkmask, fcsccan$eig2, FALSE  )
plot( thkmask, ccaeseg, axis=3, nslices=16 )
```

## Quick CCA-based prediction: FA and CBF

```{r facca3cbf,eval=TRUE,echo=FALSE}
fcsccanmat=imageListToMatrix( fcsccan$eig1, famask)
fcsccanmat=fcsccanmat/rowSums(fcsccanmat)
fcsccanproj = famat[haveAllMerge,] %*% t( fcsccanmat )
fcsccanmat2=imageListToMatrix( fcsccan$eig2, thkmask)
fcsccanmat2=fcsccanmat2/rowSums(fcsccanmat2)
fcsccanproj2 = cbfmat[haveAllMerge,] %*% t( fcsccanmat2 )
ccadf=data.frame( sdemog, fcsccanproj,fcsccanproj2,
  gender=demog$Sex[haveAllMerge] )
lmdl=lm( AgeAtScan ~ ., data=ccadf[traing,] )
mdl=randomForest( AgeAtScan ~., data=ccadf[traing,] )
pviq = predict( mdl, newdata=ccadf[ testg, ] )
print( mean( abs( pviq - ccadf[testg,]$AgeAtScan) ) )
cor.test( pviq, ccadf[testg,]$AgeAtScan )
```





# Review

## Discussion

* Reviewed sparse dimensionality reduction

* These methods update traditional dimensionality reduction tools for neuroimaging

* We showed strategies for using these modalities together to find meaningful predictors.

* Performed some visualization and interpretation.
